Drosophila macrophages control systemic cytokine levels in response to oxidative stress via a non-canonical DNA damage repair signaling cascade

Version 1

Abstract

Summary

Environmental factors, infection, or injury, cause oxidative stress in diverse tissues, resulting in immune activation and loss of tissue homeostasis. Effective stress response cascades, conserved from invertebrates to mammals, ensure reestablishment of homeostasis and tissue repair. Plasmatocytes, the Drosophila macrophage-like cells, are thought to respond to oxidative stress by immune activation, however the signaling cascades involved in oxidative stress sensing and subsequent immune activation are yet to be defined. Furthermore, their role in modulating and controlling oxidative stress response to facilitate tissue repair and survival of the organism is not resolved. Here we describe the responses of hemocytes in adult Drosophila to oxidative stress and the essential role of non-canonical DNA damage repair activity in direct “responder” hemocytes to control JNK-mediated stress signaling, systemic levels of the cytokine upd3 and subsequently susceptibility to oxidative stress. Our results point to an essential systemic role of hemocytes in controlling systemic oxidative stress response in Drosophila, including energy mobilization for potential tissue repair.

Introduction

Metazoan organisms must constantly adapt to the surrounding environment. Environmental factors, including radiation, pathogens or toxins, constantly lead to the generation of reactive oxygen species (ROS)1. Low levels of ROS support cellular pathways like immune cell differentiation and even immune function such as antimicrobial defense2, 3. At high concentrations, ROS leads to oxidative stress, a deleterious process that negatively impacts protein synthesis and stability, lipids as well as overall genome stability. ROS mediated cellular damage can further result into degenerative diseases or cancer4. Therefore, organisms employ efficient repair and defense mechanisms to circumvent high levels of oxidative stress and subsequent organ damage. Hence, keeping the balance between the production and the antioxidant-dependent clearance of ROS is essential for the survival and well-being of the affected organism5. Oxidative stress induced tissue damage is most often followed by a strong local immune response. These molecular processes are conserved across vertebrates and invertebrates such as the fruit fly Drosophila melanogaster6.

Hemocytes, the Drosophila blood cells, are essential cellular components of the immune and stress response machinery. They consist of three different subtypes: crystal cells, lamellocytes and plasmatocytes7. Plasmatocytes are considered to be macrophage equivalents in Drosophila and comprise up to 95 % of all hemocytes in the adult fly8. They share many functions as well as the multilayered ontogeny with vertebrate tissue macrophages9, 10. Adult plasmatocytes are actively involved in many immune functions such as pathogen recognition, defense and phagocytosis but also chronic immune activation upon metabolic stress by dietary lipids11, 12. Recent studies highlighted the essential role of adult plasmatocytes as key players in regulation of tissue homeostasis, including the gut by regulating the intestinal stem cell niche or in muscles by controlling glucose homeostasis13, 14. Similar to higher eukaryotes, levels of ROS are used as a central antimicrobial defense mechanisms of hemocytes against invading pathogens15. Moreover, larval hemocytes respond differentially to ROS, with two cell subsets mounting a biphasic immune reaction to invading bacteria by controlling hemocyte spreading and adhesion as well as crystal cell rupture15. Larval hemocytes have been also attributed to respond to increased oxidative stress of other tissues, as seen in response to oxidative stress in the eye-antenna imaginal disc16. Similarly, by regulating the intestinal stem cell (ISC) proliferation in the damaged midgut during infection-induced or chemical-induced oxidative stress14, adult hemocytes were described to release the bone morphogenetic protein (BMP) homologue decapentaplegic (dpp) to facilitate tissue repair14. Other studies pointed to an essential role of hemocyte-derived unpaired-3 (upd3), which is the invertebrate orthologue of Interleukin-6, in the regulation of ISC proliferation during septic injury17, as well as wound healing during sterile injury18. In line with these findings, hemocytes have been implicated in organ-to-organ immune signaling for example upon infection-induced injury in the gut and antimicrobial peptide (AMP) production in the fat body19. Hemocyte activation is thus a hallmark of oxidative stress in defined tissues, though their role in modulating and controlling oxidative stress response to facilitate tissue repair and survival of the organism remains undefined.

Here, we demonstrate an essential role of hemocytes to control susceptibility to oxidative stress, where hemocytes orchestrate systemic oxidative stress response by defined cellular states identified by unbiased transcriptomic profiling. Plasmatocytes, which make up 95% of all hemocytes, segregate in oxidative stress responder subclusters characterized by immune activation and cellular stress response and potential bystander subsets more associated with metabolic gene regulations. Mechanistically, we unravel that the direct immune response to oxidative stress by plasmatocytes seems to be essentially controlled by non-canonical DNA damage signaling which subsequently regulates Jun kinase (JNK) activation and induction of the pro-inflammatory cytokine upd3. We further describe that plasmatocytes are a major cellular source of upd3 and that regulation of upd3 levels in the adult fly determine the organism’s susceptibility to oxidative stress, most likely by orchestrating systemic stress responses in other tissues, such as energy mobilization in the fat body.

Results

Hemocytes control susceptibility to oxidative stress

In vertebrates and invertebrates, prolonged oxidative stress can cause severe tissue damage, hence controlled activation of stress signaling cascades are essential for the survival of the organism and the resolution of oxidative stress with subsequent tissue repair. Yet the precise role of macrophages in the initiation and control of oxidative stress signaling needs to be determined.

To analyze the role of macrophages in signaling cascades to oxidative stress, we took advantage of the well-established Paraquat (PQ)-feeding model in adult Drosophila melanogaster to induce systemic oxidative stress20. Feeding of 7-days old w;Hml-Gal4,UAS-2xeGFP/+ (Hml/+) flies with PQ-supplemented sucrose solution led to an increased lethality of adult Drosophila after 18 hours in a dose dependent manner (0mM: 1.25 ± 1.25 %, 2mM: 2.5 ± 1.6 4%, 15mM: 27.78 ± 5.47 %, and 30mM: 38.75 ± 6.67 %) (Figure S1A). In agreement with previous studies20, 15mM PQ was chosen for further experiments (Figure 1A). First, we determined which systemic changes are associated with susceptibility to oxidative stress. Infection induced damage in enterocytes can result in gut leakage and subsequent death17. We tested if the death of flies after 18 hours on 15mM PQ food was associated with PQ-induced damage of gut enterocytes and resulting gut leakage, but we found no indication of gut leakage on 15mM PQ food (Figure S1B). Only very few flies on control food and PQ indicated gut leakage of the food color to thorax and abdomen which was unrelated to the treatment and not altered between the groups (control: 3.75%, 2mM PQ: 2.53 %, 15mM PQ: 1.11 % and 30mM PQ: 3.75 %) (Figure S1B). We investigated if aberrant systemic induction of immune signaling cascades are seen upon PQ-mediated oxidative stress. For this gene expression levels of different known immune activated genes were analyzed by qPCR such as unpaired (upd) cytokines (upd1, upd2, upd3), Jak/STAT down-stream targets (Turandot-A (TotA), suppressor of cytokine signaling at 36E (Socs36E)), stress induced JNK activation (puckered (puc)) (Figure 1B), as well as expression of nuclear factor ‘kappa-light-chain-enhancer’ of activated B-cells (Nfkb)-induced antimicrobial peptides (AMP) and further cytokines such as eiger, dpp and dawdle (daw) (Figure S1C-D). These analyses did not reveal significant differences for these gene groups on a systemic level. In contrast, genes associated with carbohydrate metabolism such as insulin receptor (InR), phosphoenolpyruvate carboxykinase (Pepck), and Thor were significantly upregulated upon PQ-feeding (Figure 1C). These findings point to an activation of forkhead box, sub-group O (foxo)-target genes and a potentially reduced insulin signaling, despite the lack of significant alterations in the expression of the Insulin-like peptides (Ilps) Ilp2, Ilp3 and Ilp5 (Figure 1C).

Loss of hemocytes increases the susceptibility of adult Drosophila to oxidative stress by Paraquat feeding.

(A) Survival of Hml/+ flies with control food (n=20) and 15mM Paraquat (n=28) at 29°C. Five independent experiments were performed. Student’s unpaired t-test: ****p<0.0001, each data point represents a sample with 10 flies. (B-C) RTqPCR of whole flies were performed to investigate gene-expressions of the JAK-STAT-signaling pathway (B) and insulin-signaling pathway (C). All transcript levels were normalized to the expression of the loading control Rpl1 and are shown in arbitrary units [au]. Each data point (n=5) represents a sample of three whole flies. Student’s unpaired t-tests were performed for each transcript to compare the gene-expression between control and treated flies: **p<0.01; ***p<0.001; ****p<0.0001. (D) Glucose and (E) glycogen levels of whole flies were measured in control and PQ-treated flies. Each dot (n=3) represents a sample containing three flies. Student’s unpaired t-test: **p<0.01. (F) Triglyceride (TG) amounts were determined via thin-layer chromatography. Student’s unpaired t-test: **p<0.01. (G) Representative images of 7 day old HmlΔ4>eGFP flies treated with 5% sucrose solution and 15mM PQ respectively. Scale bars = 500µm. (H) Hemocyte quantification of 7 day old HmlΔ4>eGFP flies treated with control food (5% sucrose solution) (n=5) or 15mM Paraquat (n=5). Each data point represents one fly. Statistical significance was tested using student’s unpaired t-test: p=0.508. (I) Survival of crq-Gal80ts/+ control flies (n=13) and hemocyte depleted crq-Gal80ts>reaper flies (n=7) on control food (n=20) and 15mM Paraquat (n=28) at 29°C. Student’s unpaired t-test: ****p<0.0001, each data point represents a sample with 10 flies.

To test the consequences for energy mobilization and nutrient storage, we measured free glucose content, as well as glycogen and triglyceride levels in whole flies after 18 hrs on control or 15mM PQ. Although no changes in free glucose levels (control: 1.49±0.02µg and PQ: 1.50±0.07µg, ns p=0.954) (Figure 1D) were found, we observed a clear decrease in stored glycogen (control: 1.02±0.07µg and PQ: 0.33±0.08µg, **p=0.003) (Figure 1E) and stored triglycerides (Figure 1F) in PQ-treated flies compared to control. Next, we examined histological sections of flies to define which tissues might be affected. Hematoxylin and eosin (H&E) stainings showed a loss of stored lipids in the abdominal fat body compared to flies kept under control conditions (Figure S1E). Fat body development, energy mobilization and storage have been recently connected to the presence, cell number and subsequent cross-talk of hemocytes with fat body21, therefore we quantified GFP+ hemocytes in Hml/+ flies after 18 hrs on PQ or control food (Figure 1G-H). Our analysis did not reveal alterations in hemocyte numbers and especially no oxidative stress induced loss of hemocytes (control: 243.4±24.33 hemocytes per fly and PQ: 219.6±24.22 hemocytes per fly, ns p=0.508) (Figure 1G-H). To evaluate if hemocytes are essential contributors to the stress-mediated metabolic changes and susceptibility to oxidative stress, we used “hemocyte-deficient” crq-Gal80ts>reaper (w;tub-Gal80ts/UAS-CD8-mCherry;crq-Gal4/UAS-reaper) flies and tested their survival rate compared to control flies on PQ food (Figure 1I). Hemocyte-deficient flies displayed a significantly higher mortality rate on 15mM PQ (43.33±6.30%) compared to control flies (13.08±2.63%). Our findings show detrimental changes in energy mobilization upon oxidative stress in the absence of profound over-activation of immune signaling cascades, however the presence or absence of hemocytes seem to be a key determinant of the susceptibility to oxidative stress in adult Drosophila. This strongly points to an essential role for hemocytes in controlling stress response and organismal survival upon oxidative stress.

Paraquat exposure generates diverse transcriptional states of adult hemocytes with specific transcriptional state associated with oxidative stress response

To gain further insights into the alterations induced in the adult fly upon oxidative stress and especially to the hemocyte compartment, we decided to take advantage of unbiased single nucleus RNA-sequencing (snRNA-seq). We isolated nuclei from control food and PQ-treated Hml-DsRed.nuc flies with a modified Frankenstein protocol22 and performed fluorescence activated nuclear sorting of DAPI+Draq7+ nuclei (Figure S2A). From two independent sorts we obtained a total of 29000 nuclei which were utilized for subsequent droplet-based barcoding of single nuclei, and generation of single nucleus cDNA libraries. We could successfully obtain gene expression data from 11839 nuclei (Figure S2B-D). Quality control analysis followed by unsupervised hierarchical clustering of all analyzed nuclei using Seurat package23 revealed 20 distinct nuclei clusters (C) with differential gene expression (Figure 2A). Cell identity was annotated according to the cell type-specific signatures determined by the Fly Cell Atlas consortium (Figure 2A)24. All annotated cell types were detected in PQ-treated and control flies (Figure 2B).

Unbiased single nuclei transcriptomic profiling identified diverse transcriptomic states of hemocytes associated with oxidative stress response.

(A) Uniform Manifold Approximation and Projection (UMAP) visualization of single nucleus states from flies exposed to 15mM PQ containing or control food. Dashed lines indicate the broad cell types. Male accessory gland main cells (MAG), Malpighian tubule principal cells (MTC), male germline cells (MGC), outer photoreceptor cells (PRC), tracheolar cells (TC), epithelial cells (EC). Colors indicate distinct clusters. n=11839 nuclei are shown. (B) UMAP visualization of nuclei from A colored by treatment. Nuclei from control group are labeled in red and nuclei from PQ group are labeled in blue. Dashed lines indicate the broad cell type as A. (C) UMAP visualization of hemocytes from A after sub-setting and re-clustering. Colors and dashed lines indicate distinct clusters. n=1354 nuclei are shown. (D) UMAP visualization of hemocytes clusters colored by treatment. Nuclei from control group are labeled in red and nuclei from PQ group are labeled in blue. Dashed lines indicate distinct clusters. n=1354 nuclei are shown. (E) Heat map of 20 genes enriched in each hemocytes cluster. Gene names are indicated on the left. Colors in the heat map correspond to normalized scaled expression. (F) Volcano plots comparing pseudo bulk gene expression of individual hemocytes cluster vs all remaining hemocytes. The –log10-transformed adjusted P value (P adjusted, y-axis) is plotted against the log2-transformed fold change (FC) in expression between the indicated cluster vs remaining hemocytes (x-axis). Genes under log2 FC and – Log10 P value cut off are shown in grey (NS). Genes above log2 FC but under P value cut off in green, Genes under log2 FC and above –Log10 P value cut off are shown in blue and genes above log2 FC and – Log10 P value cut off are shown in red.

Using cell type-specific signatures, three of the clusters were annotated as hemocytes (clusters: 2, 4 and 11) (Figure 2A-B). Re-scaling and clustering of all 1354 identified hemocyte nuclei revealed 8 distinct clusters (Figure 2C). We verified hemocyte-specific signature genes across these clusters including genes such as the transcription factor Serpent (Srp) and one or more of the following hemocyte signature genes: Hemolectin (Hml), croquemort (crq), Nimrod C1 (NimC1), eater, and Peroxidasin (Pxn) (Figure S3A-L)25, 26. Cluster 8 expressed various crystal cell-specific markers including lozenge (lz), pebbled (peb), prophenoloxidase 1 (PPO1), and prophenoloxidase 2 (PPO2), whereas all other clusters seemed to be plasmatocytes (Figure S3A-L). In line with previous reports, the embryonic and larval hemocyte markers Hemese (He) (data not shown) and singed (sn) were not expressed across any of the identified clusters (Figure S3H). Interestingly, when we defined which hemocyte clusters were enriched in control flies or PQ-treated flies, the two plasmatocyte clusters C1 and C7 were primarily derived from control flies, whereas C2, C3, C4 and C6 are predominantly found in PQ-treated flies (Figure 2D and Figure S3M). The plasmatocyte cluster C5 as well as the crystal cell cluster C8 presented in both conditions to similar proportions (Figure 2D and Figure S3M).

Next, we analyzed the TOP20 differentially expressed (DE) genes across all hemocyte clusters. We found clear differences between the 7 distinct plasmatocyte clusters and the crystal cell cluster (Figure 2E and Table S1). The crystal cell cluster C8 showed a high expression of crystal cell specific genes including lz, peb, PPO1 and PPO2 (Figure 2E). Plasmatocyte cluster C1, which was mostly present in control conditions, had a remarkable expression of hemocyte specific marker genes such as Hml and Pxn, as well as migration-associated genes including the semaphorins Sema2a and Sema5c (Figure 2E). Yet this cluster also showed higher expression of genes contributing to metabolic regulations such as the L-gulonate 3-dehydrogenase Had2 and the gluconolactonase regucalcin (Figure 2E). The cluster C7 and C1, mostly found in control flies, substantially shared gene expression profiles. In contrast, cluster C2 and C3 were nearly exclusively found in PQ-treated flies and showed expression of genes involved in metabolic adaptations, with high expression of the glucose transporter 4 enhancer factor (Glut4EF) or the phosphatidate phosphatase Lipin (Lpin) in C2 or the induced expression of the lipase brummer (bmm), the autophagy-associated gene Atg17 or the lipophorin receptor LpR1, involved in uptake of neutral lipids from the hemolymph, in C3 (Figure 2E). Overall gene expression profiles in clusters C2 and C3 showed an induction of insulin signaling, increased autophagy and lipolysis pointing to adaptation of energy mobilization in these clusters upon PQ treatment (Figure 2E). Cluster C4 was characterized by a remarkable upregulation of genes involved in translation including many ribosomal proteins such as RpL8, RpL31 or RpS13 (Figure 2E). Furthermore, C4 included genes associated with unsaturated fatty acid metabolism such as Desaturase 1 (Desat1). Similarly, cluster C5 also showed induction of genes involved in fatty acid metabolism such as ELOVL fatty acid elongase 7 (ELOVL) (Figure 2E). Of note, we also found Tie-like receptor tyrosine kinase (Tie) specifically expressed in this cluster, which is described to bind PDGF- and VEGF-related factor (Pvf) ligands and contribute to cell survival and migration, as well as actin-associated binding proteins including polychaetoid (pyd) and formin 3 (form3). All these gene expression changes are pointing to a plasmatocyte cluster with a gene expression profile associated with mobility and migration. Among all plasmatocyte clusters, we found cluster C6 which presented a strong immune activation phenotype with induction of Jak/STAT target genes including Fasciclin 3 (Fas3) and Turandot-M (TotM) among its TOP20 DE genes, but also activation of oxidative stress response genes such as the heat shock proteins Hsp23, Hsp26 and Hsp27 or cytoskeletal genes including myosin Myo31DF or spire (spir) (Figure 2E).

Due to the partially shared expression profile of genes across hemocyte clusters, we wondered if there were genes exclusively expressed in individual clusters. However, cluster C1 showed that Sema2a seems to be exceptionally high expressed in these homeostatic plasmatocytes and genes associated with metabolic regulation (e.g. InR, Lpin or bmm) or Jak/STAT activation (e.g. Fas3), as seen in PQ-associated clusters, are downregulated (Figure 2F). Cluster C5 showed enrichment for genes such as the transcription factor grainy head (grh) but also cytoskeletal-associated genes such as jitterbug (jbug) and form3, suggesting a migratory potential of plasmatocytes from this cluster within their host tissues (Figure 2F). Interestingly, plasmatocyte cluster C6 presented with the most profound gene expression changes compared to all other clusters. Similar to our previous observations, we identified that genes specifically expressed in this cluster were associated with Jak/STAT pathway activation (e.g. Fas3, TotM), cytoskeletal rearrangements and migration (e.g. Myo31DF, spir or cheerio (cher)) and response to oxidative stress (e.g. Nicotinamide amidase (Naam) or Hsp23) (Figure 2F). We also included the crystal cell cluster C8 in this analysis and verified unique expression of crystal cell markers compared to the seven plasmatocyte clusters (Figure 2F). Our snRNA-seq analysis revealed that adult hemocytes respond differentially to oxidative stress with a specific cluster of plasmatocytes associated with immune activation and stress response. Despite our findings that on systemic level we did not detect a strong induction of immune activation upon oxidative stress, we now provide evidence that plasmatocyte cluster C6 showed a substantial immune response towards oxidative stress.

Direct response to oxidative stress in vitro mimics the transcriptional changes seen in cluster 6 plasmatocytes in vivo

Plasmatocytes responded to PQ-mediated oxidative stress in different cellular states. We defined cluster C6 as a cell state which is associated with oxidative stress response and immune activation. Since cluster C6 showed gene expression associated with oxidative stress and immune activation we focused our further analysis on this cluster. To further substantiate our findings and address the signaling pathways that could lead to diverse transcriptional states, we inferred the transcription factor (TF) network based on gene expression data (Figure 3A). This analysis revealed that genes with binding sites for Xbp1 were enriched. Xbp1 is a transcription factor associated with stress response. In addition to this, the Nf-κB-related transcription factor Relish (Rel) and kayak (kay), a component of the activator protein-1 (AP-1) transcription factor complex induced by JNK signaling, were also enriched in this cluster (Figure 3A). To determine if the gene expression signature of cluster C6 plasmatocytes reflects a functional state of plasmatocytes directly induced by oxidative stress, we treated S2 cells, a Drosophila plasmatocyte cell line, with different concentrations of PQ (15mM and 30mM) (Figure S4). First, we performed flow cytometric analysis of ROS levels via CellROX staining after 6 and 24 hours PQ treatment compared to controls (Figure S4A-B). We observed a dose-dependent increase in intracellular ROS at all time points analyzed with higher ROS levels at 6 hrs (0mM: 1903±64 median fluorescence intensity (MFI), 15mM: 6735±1096 MFI, and 30mM: 7435±237 MFI) compared to 24 hrs (0mM: 1236±135 MFI, 15mM: 4456±228 MFI, and 30mM: 5557±303 MFI) (Figure 4C). Increase in ROS and oxidative stress can lead to cellular damage such as damage in nuclear and mitochondrial DNA.

A specific cluster of plasmatocytes responds to oxidative stress with immune activation by Jak/STAT, DDR and JNK signaling.

(A) Key transcription factors (TFs) regulating hemocytes states. Heat map of scaled TF activity at a single cell level in hemocytes states from Figure 2C. Colors in the heat map correspond to scaled expression. Numbers on top indicate the cluster identity. Labels on the left and right indicate the TFs. (B) Feature plots showing scaled expression of selected genes associated with JAK/STAT signaling (upd3, Socs36E, TotA, TotM, Fas3), JNK and DNA damage signaling (puc, Gadd45), insulin signaling (Thor, InR), TGFβ signaling (dpp, daw) and TNF signaling (egr).

Loss of non-canonical DNA damage signaling activity in hemocytes leads to an increase in systemic upd3 levels and a higher susceptibility to oxidative stress.

(A) Survival of Hml/+ (n= 28), Hml>mei41-IR (n= 23); Hml>tefu-IR (n= 24); Hml>mei41-IR,tefu-IR (n= 16) and Hml>nbs-IR (n= 22) flies on control food and PQ. Each data point represents a sample of 10 flies. Mean ± SEM is shown. Three independent experiments were performed. One-way ANOVA: ***p<0.001; ****p<0.0001. (B) Hemocyte quantification of Hml/+ control flies and DDR-knockdowns on control food (C) and 15mM Paraquat (P). Each data point represents one fly (n=5 for all groups). Mean ± SEM is shown. One way ANOVA: Hml/+ (C) vs Hml>tefu-IR (C), **p=0.0052; Hml>tefu-IR (C) vs Hml>tefu-IR (P), *p=0.03 (C) Comet assay of isolated hemocytes from Hml/+ (C: n=73; P: n=129), Hml>mei41-IR (C: n=119; P: n=132) and Hml>tefu-IR (C: n=175; P: n=125) - flies on control food and PQ. Olive tail moment of comet assays of sorted hemocyte nuclei is shown. One-way ANOVA: *p<0.05; ***p<0.001; ****p<0.0001. (D) Gene expression analysis for Jak/STAT target genes via RT-qPCR in Hml/+ control flies and DDR-knockdowns on control food (C) and 15mM Paraquat (P). All transcript levels were normalized to the expression of the loading control Rpl1 and are shown in arbitrary units [au]. Each data point (n=5) represents a sample of three individual flies. Mean ± SEM is shown. Two-way ANOVA: *p<0.05; **p<0.01; *** p<0.001; ****p<0.0001. (E) Gene expression analysis for insulin signaling target genes and Ilps via RT-qPCR in Hml/+ control flies and DDR-knockdowns on control food (C) and 15mM Paraquat (P). All transcript levels were normalized to the expression of the loading control Rpl1 and are shown in arbitrary units [au]. Each data point (n=5) represents a sample of three individual flies. Mean ± SEM is shown. Two-way ANOVA: *p<0.05; **p<0.01; *** p<0.001; ****p<0.0001. (F) Triglyceride (TG) levels of Hml/+, Hml>mei41-IR, Hml>tefu-IR, Hml>mei41-IR,tefu-IR and Hml>nbs-IR flies on control food and PQ determined via thin-layer chromatography (TLC). One representative TLC is shown (left panel). Quantification thereof is shown on the right. Each data point (n=4 per group) represents a sample of ten individual flies. Mean ± SEM is shown. n.d. = not detectable. One-way ANOVA was performed for both groups (15mM PQ and control) separately: *p=0.0106.

We next tested whether, we now aimed to test if plasmatocytes directly exposed to oxidative stress might also experience these type of stress response by performing comet assays of treated and untreated S2 cells. Upon PQ treatment, S2 cells showed a remarkable and dose-dependent increase in DNA damage compared to untreated cells (0mM: 101.0±2.1, 5mM: 117.6±3.3, 10mM: 136.6±3.8, and 15mM: 147.8±5.5, data shown as olive tail moment) (Figure S4D). We were then wondering if PQ-treated S2 cells would resemble cluster C6 plasmatocytes in vivo and performed gene expression analysis of S2 cells treated with 15mM PQ after 6 and 24 hrs. Comparable to our results for cluster C6 in vivo, we found a profound induction of the Jak/STAT target genes by RTq-PCR. In S2 cells, we also detected a significant increase of the Jak/STAT cytokines upd1, upd2 and upd3 (Figure S4E), which we did not see before in vivo. Furthermore, we detected an increased expression of the JNK-target gene puckered (puc), but barely any induction of metabolic genes (Figure S4E). All these results further support our hypothesis that direct exposure of S2 cells with PQ in vitro mimics gene expression changes seen in C6.

Because we found enrichment of genes associated with the Nf-κB-related transcription factor Rel in cluster C6, we further verified if any other cytokines or AMPs are induced in PQ-treated S2 cells. We identified a slight but significant induction of the cytokine daw, but no other alterations in dpp or the AMPs Drs, Dro or Mtk (Figure S4F). The profound induction of Jak/STAT target genes together with the activation of JNK target genes and oxidative stress induced DNA damage in S2 cells in vitro highly support our hypothesis that cluster C6 plasmatocytes reflect a cellular state associated with the direct response to oxidative stress in vivo. Along this line, we went back to our snRNA-seq data and specifically searched for Jak/STAT target genes (upd3, Socs36E, TotA, TotM and Fas3) across all clusters (Figure 3B). Here we identified an enriched expression of these genes in cluster C6, and a unique expression of the pro-inflammatory cytokine upd3 in cluster C6 which was not detected in any other hemocyte cluster (Figure 3B). Interestingly, we also identified that Growth arrest and DNA damage-inducible 45 (Gadd45), a gene linking DNA damage signaling and stress induced JNK signaling, was enriched in cluster C6 (Figure 3B). This could indicate that cluster C6 also undergoes oxidative stress mediated DNA damage with subsequent JNK activation, similar to what we have seen in S2 cells in vitro. Of note, the JNK target gene puc was found to be highly expressed across all hemocyte clusters analyzed in vivo (Figure 3B). We looked again into the expression of metabolic genes including Thor and InR as well as other cytokines such as dpp, daw or egr but could not detect a specific enrichment in cluster C6 similar to our results from PQ-treated S2 cells in vitro (Figure 3B). These findings from the snRNA-seq analysis in vivo and S2 cells in vitro support the hypothesis that cluster C6 plasmatocytes represent a functional state of activated plasmatocytes directly responding to oxidative stress by JNK signaling and Jak/STAT pathway activation. Furthermore, we found evidence that cluster C6 specifically expresses the cytokine upd3, which could reflect an important mediator for oxidative stress response.

Loss of canonical DNA damage signaling in adult hemocytes results in elevated upd3 levels and a decreased survival on oxidative stress

We observed that hemocytes respond with a distinct gene expression pattern specific to oxidative stress (Figure 1-3). Previously it has been observed that oxidative stress-induced DNA damage supports immune activation in macrophages via a non-canonical DNA damage response27. Based on our findings, we predict that also in adult hemocytes oxidative stress not only drives genome instability and activates repair, but is rather essential to induce immune activation via specific gene expression changes in cluster C6 plasmatocytes. To test if DNA damage signaling is involved in the response of adult hemocytes to oxidative stress, we analyzed flies with a hemocyte-specific knock-down for different genes of the DNA damage repair (DDR) machinery including the DNA damage sensing kinases telomere fusion (tefu) and meiotic 41 (mei-41) and the MRN complex protein nbs. Of note, by targeting adult hemocytes, mostly plasmatocytes are targeted which make up 95% of all hemocytes in the adult fly.

We generated Hml>mei-41-IR (w;Hml-Gal4,UAS-2xeGFP/UAS-mei-41-IR), Hml>tefu-IR (w;Hml-Gal4,UAS-2xeGFP/UAS-tefu-IR), Hml>mei41-IR;tefu-IR (w;Hml-Gal4,UAS-2xeGFP/UAS-mei-41-IR;UAS-tefu-IR) and Hml>nbs-IR (w;Hml-Gal4,UAS-2xeGFP/UAS-nbs-IR) flies (Figure S5A). First, we compared their susceptibility to PQ treatment after 18 hrs and found significantly increased susceptibility to the treatment across all DDR-deficient lines compared to the control genotype (Hml/+: 10.7±1.2%, Hml>mei-41-IR: 40.7±4.8%, Hml>tefu-IR: 20.7±4.3%, Hml>mei41-IR;tefu-IR: 37.5±4.8% and Hml>nbs-IR: 48.2±6.7%) (Figure 4A). In these flies, decreased survival after PQ did not correlate to an overall reduction in lifespan on normal food and a decrease in fitness (Figure S5B).

Next, we inquired if DDR-deficiency in hemocytes might increase susceptibility of the fly to other stresses. For this we challenged all genotypes with starvation. Importantly, we did not observe a decreased survival of DDR-deficient flies on starvation compared to the control genotype, indicating that the observed susceptibility to oxidative stress of flies with a DDR-deficiency in hemocytes is specifically caused by oxidative stress due to PQ treatment (Figure S5C). As defects in the DDR machinery in hemocytes might already affect the development during embryonic and larval stages, we used a hemocyte-specific temperature-inducible knock-down of mei-41, tefu and nbs1 which was induced by a temperature shift to 29°C directly after eclosion from the pupae to induce a knockdown in adult hemocytes only. We analyzed the lifespan of Hml-Gal80ts>mei-41-IR (w;Hml-Gal4,UAS-2xeGFP/UAS-mei-41-IR;tub-Gal80), Hml-Gal80ts>tefu-IR (w;Hml-Gal4,UAS-2xeGFP/UAS-tefu-IR;tub-Gal80), Hml-Gal80ts>mei-41-IR;tefu-IR (w;Hml-Gal4,UAS-2xeGFP/UAS-mei-41-IR;tub-Gal80/UAS-tefu-IR) and Hml-Gal80ts>nbs-IR (w;Hml-Gal4,UAS-2xeGFP/UAS-nbs-IR;tub-Gal80) compared to control flies Hml-Gal80ts/+ (w;Hml-Gal4,UAS-2xeGFP/+;tub-Gal80) and again found no alterations in overall lifespan of all genotypes analyzed on control food except a slight decrease in lifespan of Hml-Gal80ts>mei-41-IR;tefu-IR. (Figure S5D). In contrast, we again found a higher susceptibility to oxidative stress via PQ-feeding in the DDR-deficient lines compared to control (Hml-Gal80ts/+: 30.3±5.2%, Hml-Gal80ts>mei-41-IR: 36.2±4.3%, Hml-Gal80ts>tefu-IR: 52.4±6.9%, Hml-Gal80ts>mei-41-IR;tefu-IR: 72.3±6.3% and Hml Hml-Gal80ts>nbs-IR: 36.2±4.3%) (Figure S5E), which directly confirms our previous results and conclusions (Figure 4A).

As we have seen before that loss of hemocytes is detrimental for the survival on oxidative stress, we wanted to exclude a reduction in hemocytes in the DDR deficient fly lines as a cause of the higher susceptibility to oxidative stress. We quantified GFP+ hemocytes in Hml>mei-41-IR, Hml>tefu-IR, Hml>mei-41-IR;tefu-IR and Hml>nbs-IR flies (Figure 4B, Figure S5A). We did not find alteration in hemocyte number across all fly lines on control food (with the exception of an increase in hemocytes in the Hml>tefu-IR line), nor did we observe any effect of PQ treatment on the number of hemocytes in DDR-deficient flies (Figure 4B, Figure S5A).

Next we analyzed if the loss of the DDR signaling machinery via knockdown of tefu and mei41 results in increased DNA damage in hemocytes. Comet assays of isolated hemocytes from DDR deficient flies and the control genotype on PQ and control food only indicated a slight increase in DNA damage in these flies, which demonstrates that the higher susceptibility of DDR deficient flies is not due to a loss of hemocyte numbers or a substantial increase in DNA damage, but rather due to changes in the downstream DDR-signaling events modulating immune activation of hemocytes (Figure 4C). Other studies demonstrated that upon tissue damage JNK-mediated secretion of the JAK/STAT cytokines upd1, upd2 and upd3 is a critical step for innate immune signaling28, 29. Furthermore, DNA damage signaling is essential to induce such an immune response and to regulate insulin signaling30. To elucidate the implication of DDR signaling on hemocyte activation and oxidative stress resistance, we measured the gene expression levels of upd cytokines as well as JNK and Jak/STAT targets in DDR deficient flies and controls on PQ and control food. Loss of DNA damage signaling in hemocytes induced a significant induction of upd3 expression upon PQ treatment in Hml>tefu-IR and Hml>mei-41-IR;tefu-IR flies (Figure 4D). In contrary, upd1 and upd2 were not significantly changed in DDR-deficient flies (Figure 4D). We measured the Jak/STAT target genes Socs36E and TotA on systemic level. A slight induction of Socs36E expression in all DDR-deficient genotypes on PQ but only Hml>mei-41-IR;tefu-IR flies showed a significant induction of Socs36e upon PQ treatment compared to control food and Hml/+ flies on PQ (Figure 4D). TotA expression tended to be reduced upon PQ treatment across all genotypes analyzed. However, when we checked the levels of puc as a JNK signaling target gene, we found a significant induction on PQ treatment in Hml>tefu-IR, Hml>mei-41-IR;tefu-IR and Hml>nbs-IR flies, which was not seen in control flies upon PQ treatment (Figure 4D). When we analyzed the gene expression of genes involved in metabolism and energy mobilization, we detected a significant induction and dysregulation of Thor and InR in all DDR-deficient genotypes upon PQ treatment compared to control flies (Figure 4E), as well as a slight induction of Pepck1 (Figure 4E). To exclude that these changes in energy mobilization are due to altered expression of Ilps, we analyzed the levels of Ilp2, Ilp3 and Ilp5 in all genotypes on control and PQ-food, but could not detect remarkable alterations here (Figure 4E). To further exclude that the higher susceptibility of DDR-deficient flies and seen alterations in stress response are due to inefficient mobilization of triglycerides, we measured stored triglycerides across all genotypes and found efficient mobilization of triglyceride storage upon PQ treatment in all genotypes compared to control food (Figure 4F).

From our results we can conclude that DDR signaling in adult hemocytes, mostly consisting of plasmatocytes, essentially determines susceptibility to oxidative stress. Loss of canonical DNA damage signaling in hemocytes, which normally would induce DDR or cell death, triggers enhanced non-canonical DNA damage signaling such as upregulation of JNK signaling and elevated expression of the pro-inflammatory cytokine upd3, as well as transcriptional alterations in energy mobilization and insulin signaling. Based on our previous results on the role of hemocytes during oxidative stress, we can therefore assume that this JNK activation and upd3 induction is essentially mediated by hemocytes.

Adult hemocytes control susceptibility to oxidative stress by JNK signaling and upd3 release

To this point, our data suggests that oxidative stress-induced DDR in hemocytes limits immune activation and stress response upon oxidative stress. In other words, enhanced DNA damage boosts immune activation and therefore susceptibility to oxidative stress. As we found that increased upd3 levels correlate with a higher susceptibility to oxidative stress of flies with DDR-deficient hemocytes, we aimed to test if hemocyte-derived upd3 is a major source to regulate susceptibility to PQ-induced oxidative stress. To do so, we either over-expressed upd3 in hemocytes in Hml>upd3 (w;Hml-Gal4,UAS-2xeGFP/UAS-upd3) flies or performed a hemocyte-specific knockdown of upd3 in Hml>upd3-IR (w;Hml-Gal4,UAS-2xeGFP/UAS-upd3-IR) flies. Both genotypes showed a normal overall lifespan compared to the Hml/+ control (Figure S5F), but when we compared their susceptibility to PQ-induced oxidative stress, Hml>upd3 flies showed a drastically decreased survival on PQ, whereas Hml>upd3-IR flies did not show a changed survival on PQ treatment after 18 hours (Figure 5A). We tested upd3 expression levels on a whole fly level and found a slight but not significant induction in Hml/+ flies after PQ treatment compared to control food, in line with our previous results (Figure 1B and 5B). As expected, Hml>upd3 flies showed increased levels of upd3 in both treatment groups compared to the other genotypes, but also showed a significant increase of upd3 upon PQ compared to control food (Figure 5B). Surprisingly, Hml>upd3-IR flies still showed a slight induction of upd3 upon PQ treatment, which could potentially be derived from another cellular source after eliminating hemocyte-derived upd3. To identify if the higher susceptibility of Hml>upd3 flies to oxidative stress was due to reduced hemocyte numbers, we quantified hemocytes across all three genotypes, but did not see any significant changes of hemocyte number in both genotypes compared to Hml/+ control flies on control food and PQ (Figure 5C and Figure S5G). Only in the Hml>upd3-IR flies we found a reduction of hemocytes between flies on control food and PQ, but without consequences for their susceptibility to oxidative stress (Figure 5C and Figure S5G). Our data indicate that elevation of hemocyte-derived upd3 is sufficient to increase the mortality of adult flies on PQ-mediated oxidative stress. Yet we could not exclude other cellular sources for upd3 upon oxidative stress other than hemocytes. Therefore, we wanted to probe if hemocyte-derived upd3 is sufficient to increase the susceptibility to PQ treatment in the absence of other cellular sources. We backcrossed Hml>upd3 flies to upd3-deficient flies (upd3null) to generate upd3null;Hml>upd3 flies (upd3null;Hml-Gal4,UAS-2xeGFP/UAS-upd3) and compared their survival on PQ to different control genotypes (w1118 and Hml/+) and upd3null flies (Figure 5D). We did not identify any differences in the susceptibility of w1118, Hml/+ and upd3nullcontrol flies to PQ treatment, however we found a significantly decreased survival in upd3null;Hml>upd3 flies when hemocytes are overexpressing upd3 (Figure 5D).

Hemocyte-derived upd3 controls susceptibility to oxidative stress in adult Drosophila.

(A) Survival of Hml/+ (n=28), Hml>upd3-IR (n=27) and Hml>upd3 (n=26) flies on 15mM PQ food. Each data point represents a sample of 10-15 flies. Mean ± SEM is shown. Three independent experiments were performed. One-way ANOVA: ****p<0.0001. (B) Gene expression analysis for upd3 via RT-qPCR in Hml/+, Hml>upd3-IR and Hml>upd3 flies on control food (-) and 15mM Paraquat (+). Each data point (n=5-6) represents a sample of three individual flies. Mean ± SEM is shown. Two-way ANOVA: ****p<0.0001. (C) Hemocyte quantifications in Hml/+, Hml>upd3-IR and Hml>upd3 flies on control food and 15mM PQ food. Each data point represents one fly (n= 5-6). Mean ± SEM is shown. One-way ANOVA: **p<0.01. (D) Survival of w1118 (n=17), Hml/+ (n=16), upd3null (n=11) and upd3null;Hml>upd3 flies (n=16) on 15mM PQ food at 29°C. Each data point represents a sample with 10-15 flies. Mean ± SEM is shown. One-way ANOVA: ****p<0.0001. (E) Survival of Hml/+ (n=78), Hml>hep[act] (n=38) and Hml>bsk-DN (n=80) flies on 15mM PQ food. Each data point represents a sample with 10 flies. Mean ± SEM is shown. One-way ANOVA: **p<0.01; ****p<0.0001. (F) Gene expression analysis for upd3 via RT-qPCR of Hml/+, Hml>hep[act] and Hml>bsk-DN flies on control food (-) and 15mM Paraquat (+). Each data point (n=5-6) represents a sample of three individual flies. Mean ± SEM is shown. Two-way ANOVA: *p<0.05.

Previous studies highlighted that JNK signaling can induce upd3 expression in hemocytes12. In line with these reports, we found enrichment of genes with kayak/AP-1 binding motifs in the plasmatocyte subset C6 which showed further upregulation of Jak/STAT target genes and also upd3 expression (Figure 3). Therefore, we wanted to verify if JNK activation in hemocytes acts as a direct downstream signal activated by non-canonical DNA damage signaling upon oxidative stress and subsequently induces the release of the pro-inflammatory cytokine upd3. We either overexpressed a constitutive active form of JNKK hemipterous (hep) in hemocytes using Hml>hep[act] (w;Hml-Gal4,UAS-2xeGFP/UAS-hep[act]) flies or overexpressed a dominant-negative version of the Jun kinase basket (bsk) in hemocytes using Hml>bsk-DN (w;Hml-Gal4,UAS-2xeGFP/UAS-bsk-DN). We compared their survival on PQ after 18 hrs and in line with our previous results, found an increased susceptibility of Hml>hep[act] flies to PQ treatment, but not of Hml>bsk-DN flies compared to the control genotype (Figure 5E). We again analyzed the overall lifespan of these flies on normal food and could not find differences between the genotypes (Figure S5H). Furthermore, when we analyzed upd3 expression in these flies upon PQ treatment, we found a significant induction of upd3 expression after PQ treatment in Hml>hep[act] compared to control food, but only a slight increase in Hml/+ and Hml>bsk-DN flies upon PQ feeding. These findings support a substantial role of JNK-mediated induction of upd3 in hemocytes upon oxidative stress. In summary, our data discovers a defined immune-responsive subset of plasmatocytes during oxidative stress. We show that this oxidative stress-mediated immune activation seems to be controlled by non-canonical DNA damage signaling resulting in JNK activation and subsequent upd3 expression, which can render the adult fly more susceptible to oxidative stress when it is over-activated.

Discussion

Similar to vertebrate macrophages, oxidative stress triggers immune activation in Drosophila hemocytes18. However this immune activation was not obvious on a systemic level upon PQ-mediated oxidative stress. In contrast, localized oxidative stress-induced immune activation in damaged tissues is an essential step of the Drosophila immune response to infection or injury19, 31. However, the precise contribution of Drosophila macrophages in this immune activation, in the organ-to-organ communication during stress response and the implication in the organism’s survival upon increased oxidative stress remained so far ill-defined. Overall, we describe responses of adult hemocytes, namely plasmatocytes, to oxidative stress with a particular subset of plasmatocytes (cluster C6) which directly responds to oxidative stress by immune activation. We demonstrated that this immune activation seems to be at least partially controlled by non-canonical DNA damage signaling in hemocytes. Loss of nbs, tefu or mei-41 in hemocytes renders the organism more susceptible to oxidative stress without signficantly affecting oxidative stress mediated DNA damage and its repair, but with an enhanced induction of JAK/STAT and JNK signaling, accompanied by changes in genes associated with energy mobilization and storage. Finally, we showed that the control of hemocyte-derived upd3 is key for the survival of the organism upon oxidative stress and that induction of the JNK/upd3 signaling axis in hemocytes is sufficient to increase the susceptibility to oxidative stress. Hence, we postulate that hemocyte-derived upd3, most likely released by the activated plasmatocyte cluster C6 during oxidative stress in vivo and subsequently controlling energy mobilization and subsequent tissue wasting upon oxidative stress.

Drosophila hemocytes have been described before to react to PQ-mediated or infection-mediated intestinal damage via release of cytokines to promote ISC proliferation in the gut14, 17. PQ is a widely used model in Drosophila to induce oxidative stress via oral uptake32. In many reports the gut has been described to be the most affected organ and ISC proliferation is an essential step to regenerate the injured intestinal epithelium33, 34. PQ treatment has been described to cause gut leakage and therefore this could be a potential cause of death for the fly14, though we did not find signs of gut leakage within the 18 hours PQ treatment with a concentration of 15mM. Aside from the gut, oral PQ treatment was shown to induce oxidative stress in many organs including hemocytes35, oenocytes36 and the brain37, indicating that the induced oxidative stress is not only localized to the gut tissue. PQ-induced oxidative stress was described before to cause neurodegeneration with parkinsonism-like symptoms, accompanied by strong immune activation and upregulation of JNK signaling in the CNS37, an activation profile similar to the identified activated plasmatocyte cluster C6 in our snRNA-seq analysis. The snRNA-seq analysis further revealed distinct hemocyte clusters with seven plasmatocyte clusters and one crystal cell cluster. Whereas the plasmatocyte clusters C1 and C7 seem to represent cellular states found in steady state conditions only, we identified several other clusters which seemed to be exclusive or at least enriched under PQ treatment. Interestingly, we only identified one cluster, C6, which specifically responded to oxidative stress via immune activation and upregulation of oxidative stress response genes but also Jak/STAT and JNK target genes. We postulate that these cells might be directly exposed to the oxidative stress because their signature resembles the one of S2 cells directly treated with PQ in vitro. Even though oxidative DNA damage in hemocytes was followed by immune activation in vitro and in vivo, it remains to be determined if this activation is induced by cytokines derived from damaged organs during oxidative stress or by the oxidative stress directly. The hypothesis that this cluster is directly exposed to oxidative stress is further supported by an upregulation of genes such as Hsps in this cluster which are known to be directly induced by oxidative stress38. Furthermore, we found evidence that oxidative DNA damage and induced non-canonical DNA damage signaling induce several of these immune activation cascades in cluster C6 plasmatocytes. This was supported by the increased DNA damage in S2 cells by PQ treatment but also the enrichment of DDR-associated genes such as Gadd45 in cluster C6, which is associated with JNK activation and is further implicated in oxidative stress induced DNA damage37. Constant DNA damage signaling and repair is an essential step in all tissues to maintain tissue homeostasis39.

In our study, we point to a potential non-canonical role of DNA damage sensing to trigger immune activation and response to oxidative stress in hemocytes. We did not find indications of canonical DNA damage response, such as induced apoptosis in hemocytes due to the oxidative DNA damage. Non-canonical DNA damage sensing has been implicated in immune responses in vertebrates and invertebrates30, 37, 40, 41. It was demonstrated that vertebrate macrophages in vitro respond similarly to immune stimuli such as infection by synergistic activation of DDR and type I interferon signaling to induce downstream immune signaling cascades27. Here, we provide new evidence in Drosophila that non-canonical DNA damage is used by immune cells in vivo to modulate and especially control their immune response including pro-inflammatory cytokine release upon oxidative stress. Localized DNA damage in larval epithelial cells has been shown to induce non-canonical DNA damage activity leading to the secretion of upds and the regulation of hemocyte expansion and activation, as well as subsequent hemocyte-mediated regulation of energy storage in the fat body30. In further support of our results, one report previously demonstrated that loss of DNA damage signaling machinery induced the release of upd3 in hemocytes14. Most likely, the identified plasmatocyte cluster C6 is a central cellular source of this signal, which needs to be further determined in future studies. Furthermore, anatomical localization of this cluster and spatial distribution of the different plasmatocyte clusters would be of interest. We describe a strong connection between the control of upd3 expression by hemocytes and the susceptibility of the adult fly to oxidative stress. Upon oxidative stress, we observe decreased triglycerides and glycogen stores and altered gene expression such as induced expression of lipolytic enzymes and other foxo target genes indicating energy mobilization from the adult fat body. Accumulating evidence throughout our study points to an implication of hemocyte-derived upd3 in the control of energy mobilization and tissue wasting during oxidative stress. These findings are in line with our previous studies where we showed that upon high fat diet hemocytes produce high levels of upd3 which induces insulin insensitivity in many tissues by activating Jak/STAT signaling in these organs and inhibiting the insulin signaling pathway12. Furthermore, we and others described a central role for hemocyte derived upds in the control of glucose metabolism in muscles during steady state and upon infection13, 42. As shown before, hemocytes produce high levels of upd3 upon infection with Mycobacterium marinum which also caused tissue wasting and extensive energy mobilization in a foxo-dependent manner11, 43.

Our study provides further evidence that hemocytes serve as essential signaling hubs in organ-to organ communication during oxidative stress, which is in line with previous immune challenge as well as steady state19, 42, 44. Our findings in Drosophila highlight the role of macrophages for the survival on oxidative stress with an urgent need for the tight control of their immune activation. We further unravel the diverse responses of hemocytes towards oxidative stress which can now serve as a baseline to further explore the spatiotemporal macrophage/hemocyte responses to oxidative stress. Future in depth analysis of the regulation of these cellular states and their coupling to tissue repair will allow to gain new insights for the treatment of diseases including cancer or neurodegenerative diseases.

eLife Assessment

This study reports important findings regarding the systemic function of hemocytes controlling whole-body responses to oxidative stress. The evidence in support of the requirement for hemocytes in oxidative stress responses as well as the hemocyte single-nuclei analyses in the presence or absence of oxidative stress are convincing. In contrast, the genetic and physiological analyses that link the non-canonical DDR pathway to upd3/JNK expression and high susceptibility, and the inferences regarding the function of hemocytes in systemic metabolic control are incomplete and would benefit from more rigorous approaches. The work will be of interest to cell and developmental biologists working on animal metabolism, immunity, or stress responses.

Reviewer #1 (Public Review):

The study examines how hemocytes control whole-body responses to oxidative stress. Using single cell sequencing they identify several transcriptionally distinct populations of hemocytes, including one subset that show altered immune and stress gene expression. They also find that knockdown of DNA Damage Response (DDR) genes in hemocytes increases expression of the immune cytokine, upd3, and that both upd3 overexpression in hemocytes and hemocyte knockdown of DDR genes leads to increased lethality upon oxidative stress.

Strengths

1, The single cell analyses provide a clear description of how oxidative stress can cause distinct transcriptional changes in different populations of hemocytes. These results add to the emerging them in the field that there functionally different subpopulations of hemocytes that can control organismal responses to stress.

2, The discovery that DDR genes are required upon oxidative stress to limit cytokine production and lethality provides interesting new insight into the DDR may play non-canonical roles in controlling organismal responses to stress.

Weaknesses

1, In some ways the authors interpretation of the data - as indicated, for example, in the title, summary and model figure - don't quite match their data. From the title and model figure, it seems that the authors suggest that the DDR pathway induces JNK and Upd3 and that the upd3 leads to tissue wasting. However, the data suggest that the DDR actually limits upd3 production and susceptibility to death as suggested by several results:

a) PQ normally doesn't induce upd3 but does lead to glycogen and TAG loss, suggesting that upd3 isn't connected to the PQ-induced wasting.

b) knockdown of DDR upregulates upd3 and leads to increased PQ-induced death. This would suggest that activation of DDR is normally required to limit, rather than serve as the trigger for upd3 production and death.

c) hemocyte knockdown of either JNK activity or upd3 doesn't affect PQ-induced death, suggesting that they don't contribute to oxidative stress-induced death. Its only when DDR is impaired (with DDR gene knockdown) that an increase in upd3 is seen (although no experiments addressed whether JNK was activated or involved in this induction of upd3), suggesting that DDR activation prevents upd3 induction upon oxidative stress.

2, The connections between DDR, JNK and upd3 aren't fully developed. The experiments show that susceptibility to oxidative stress-induced death can be caused by a) knockdown of DDR genes, b) genetic overexpression of upd3, c) genetic activation of JNK. But whether these effects are all related and reflect a linear pathway requires a little more work. For example, one prediction of the proposed model is that the increased susceptibility to oxidative stress-induced death in the hemocyte DDR gene knockdowns would be suppressed (perhaps partially) by simultaneous knockdown of upd3 and/or JNK. These types of epistasis experiments would strengthen the model and the paper.

3, The (potential) connections between DDR/JNK/UPD3 and the oxidative stress effects on depletion of nutrient (lipids and glycogen) stores was also not fully developed. However, it may be the case that, in this paper, the authors just want to speculate that the effects of hemocyte DDR/upd3 manipulation on viability upon oxidative stress involve changes in nutrient stores.

Reviewer #2 (Public Review):

Hersperger et al. investigated the importance of Drosophila immune cells, called hemocytes, in the response to oxidative stress in adult flies. They found that hemocytes are essential in this response, and using state-of-the-art single-cell transcriptomics, they identified expression changes at the level of individual hemocytes. This allowed them to cluster hemocytes into subgroups with different responses, which certainly represents very valuable work. One of the clusters appears to respond directly to oxidative stress and shows a very specific expression response that could be related to the observed systemic metabolic changes and energy mobilization. However, the association of these transcriptional changes in hemocytes with metabolic changes is not well established in this work. Using hemocyte-specific genetic manipulation, the authors convincingly show that the DNA damage response in hemocytes regulates JNK activity and subsequent expression of the JAK/STAT ligand Upd3. Silencing of the DNA damage response or excessive activation of JNK and Upd3 leads to increased susceptibility to oxidative stress. This nicely demonstrates the importance of tight control of JNK-Upd3 signaling in hemocytes during oxidative stress. However, it would have been nice to show here a link to systemic metabolic changes, as the authors conclude that it is tissue wasting caused by excessive Upd3 activation that leads to increased susceptibility, but metabolic changes were not analyzed in the manipulated flies. The overall conclusion of this work, as presented by the authors, is that Upd3 expression in hemocytes under oxidative stress leads to tissue wasting, whereas in fact it has been shown that excessive hemocyte-specific Upd3 activation leads to increased susceptibility to oxidative stress (whether due to increased tissue wasting remains a question). The DNA damage response ensures tight control of JNK-Upd3, which is important. However, what role naturally occurring Upd3 expression plays in a single hemocyte cluster during oxidative stress has not been tested. What if the energy mobilization induced by this naturally occurring Upd3 expression during oxidative stress is actually beneficial, as the authors themselves state in the abstract - for potential tissue repair? It would have been useful to clarify in the manuscript that the observed pathological effects are due to overactivation of Upd3 (an important finding), but this does not necessarily mean that the observed expression of Upd3 in one cluster of hemocytes causes the pathology.

Reviewer #3 (Public Review):

In this study, Kierdorf and colleagues investigated the function of hemocytes in oxidative stress response and found that non-canonical DNA damage response (DDR) is critical for controlling JNK activity and the expression of cytokine unpaired3. Hemocyte-mediated expression of upd3 and JNK determines the susceptibility to oxidative stress and systemic energy metabolism required for animal survival, suggesting a new role for hemocytes in the direct mediation of stress response and animal survival.

Strength of the study:

1. This study demonstrates the role of hemocytes in oxidative stress response in adults and provides novel insights into hemocytes in systemic stress response and animal homeostasis.

2. The single-cell transcriptome profiling of adult hemocytes during paraquat treatment, compared to controls, would be of broad interest to scientists in the field.

Weakness of the study:

1. The authors claim that the non-canonical DNA damage response mechanism in hemocytes controls the susceptibility of animals through JNK and upd3 expression. However, the link between DDR-JNK/upd3 in oxidative stress response is incomplete and some of the descriptions do not match their data.

2. The schematic diagram does not accurately represent the authors' findings and requires further modifications.

Author Response

The following is the authors’ response to the original reviews.

eLife assessment

This study reports important findings regarding the systemic function of hemocytes controlling whole-body responses to oxidative stress. The evidence in support of the requirement for hemocytes in oxidative stress responses as well as the hemocyte single-nuclei analyses in the presence or absence of oxidative stress are convincing. In contrast, the genetic and physiological analyses that link the non-canonical DDR pathway to upd3/JNK expression and high susceptibility, and the inferences regarding the function of hemocytes in systemic metabolic control are incomplete and would benefit from more rigorous approaches. The work will be of interest to cell and developmental biologists working on animal metabolism, immunity, or stress responses.

We would like to thank the editorial team for these positive comments on our manuscript and the constructive suggestions to improve our manuscript. We are now happy to send you our revised manuscript, which we improved according to the suggestions and valuable comments of the referees.

Public Reviews:

Reviewer #1 (Public Review):

The study examines how hemocytes control whole-body responses to oxidative stress. Using single cell sequencing they identify several transcriptionally distinct populations of hemocytes, including one subset that show altered immune and stress gene expression. They also find that knockdown of DNA Damage Response (DDR) genes in hemocytes increases expression of the immune cytokine, upd3, and that both upd3 overexpression in hemocytes and hemocyte knockdown of DDR genes leads to increased lethality upon oxidative stress.

Strengths

1. The single cell analyses provide a clear description of how oxidative stress can cause distinct transcriptional changes in different populations of hemocytes. These results add to the emerging them in the field that there functionally different subpopulations of hemocytes that can control organismal responses to stress.

1. The discovery that DDR genes are required upon oxidative stress to limit cytokine production and lethality provides interesting new insight into the DDR may play non-canonical roles in controlling organismal responses to stress.

We are grateful to referee 1 to point out the importance and novelty of our snRNA-seq data and our findings on the role of DNA damage-modulated cytokine release by hemocytes during oxidative stress. We further extended these analyses in the revised manuscript by looking deeper into the transcriptomic alterations in fat body cells upon oxidative stress (Figure 4, Figure S4). We further provide additional data to support the connection of DNA damage signaling and regulation of upd3 release from hemocytes (Figure 6F). Here we show that upd3-deficiency can abrogate the increased susceptibility of flies with mei41 and tefu knockdown in hemocytes. In line with this finding, we also show that upd3null mutants show a reduced but not abolished susceptibility to oxidative stress overall (Figure 6F), underlining the role of upd3 as a mediator of oxidative stress response.

Weaknesses

1. In some ways the authors interpretation of the data - as indicated, for example, in the title, summary and model figure - don't quite match their data. From the title and model figure, it seems that the authors suggest that the DDR pathway induces JNK and Upd3 and that the upd3 leads to tissue wasting. However, the data suggest that the DDR actually limits upd3 production and susceptibility to death as suggested by several results:

According to the referee’s suggestion, we revised the manuscript and adjusted our title, abstract and graphical summary to be more precise that DNA damage signaling seem to have a modulatory or regulatory effect on upd3 release. Furthermore, we provide now additional data to support the connection between DNA damage signaling and upd3 release. For example, we added several genetic “rescue” experiments to strengthen the epistasis that modulation of DNA damage signaling and the higher susceptibility of the fly is connected to altered upd3 levels (Figure 6F). We now provide additional data showing that the loss of upd3 rescues the susceptibility to oxidative stress in flies, which are deficient for DDR components in hemocytes.

a. PQ normally doesn't induce upd3 but does lead to glycogen and TAG loss, suggesting that upd3 isn't connected to the PQ-induced wasting.

Even though in our systemic gene expression analysis of upd3 expression, we could not detect a significant induction of upd3 upon PQ feeding. However, we found upd3 expression within our snRNAseq data in a distinct cluster of immune-activated hemocytes (Figure 3B, Cluster 6). Upon knockdown of the DNA damage signaling in hemocytes, the levels then increase to a detectable level in the whole fly. This supports our assumption that upd3 is needed upon oxidative stress to induce energy mobilization from the fat body, but needs to be tightly controlled to balance tissue wasting for energy mobilization. Furthermore, we found evidence in our new analysis of the snRNA-seq data of the fat body cells, that indeed we can find Jak/STAT activation in one cell cluster here, which could speak for an interaction of Cluster 6 hemocytes with cluster 6 fat body cells. A hypothesis we aim to explore in future studies.

b. knockdown of DDR upregulates upd3 and leads to increased PQ-induced death. This would suggest that activation of DDR is normally required to limit, rather than serve as the trigger for upd3 production and death.

Our data support the hypothesis that DDR signaling in hemocytes “modulates” upd3 levels upon oxidative stress. We now carefully revised the text and the graphical summary of the manuscript to emphasize that oxidative stress causes DNA damage, which subsequently induces the DNA damage signaling machinery. If this machinery is not sufficiently induced, for example by knockdown of tefu and mei-41, non-canonical DNA damage signaling is altered which induces JNK signaling and induces release of pro-inflammatory cytokines, including upd3. Whereas DNA damage itself is only slightly increase in the used DDR deficient lines (Figure 5C) and hemocytes do not undergo apoptosis (unaltered cell number on PQ (Figure 5B)), we conclude that loss of tefu, mei-41, or nbs1 causes dysregulation of inflammatory signaling cascades via non-canonical DNA damage signaling. However, oxidative stress itself seems to also induce upd3 release and DNA damage signaling in the same cell cluster, as shown by our snRNA-seq data (Figure 3B). Hence, we think that DNA damage signaling is needed as a rate-limiting step for upd3 release.

c. hemocyte knockdown of either JNK activity or upd3 doesn't affect PQ-induced death, suggesting that they don't contribute to oxidative stress-induced death. It’s only when DDR is impaired (with DDR gene knockdown) that an increase in upd3 is seen (although no experiments addressed whether JNK was activated or involved in this induction of upd3), suggesting that DDR activation prevents upd3 induction upon oxidative stress.

Whereas the double knockdown of upd3 or bsk and DDR genes was resulting in insufficient knockdown efficiencies, we added a rescue experiment where we combined upd3null mutants with knockdown of tefu and mei-41 in hemocytes and found a reduced susceptibility of DDR-deficient flies to oxidative stress.

1. The connections between DDR, JNK and upd3 aren't fully developed. The experiments show that susceptibility to oxidative stress-induced death can be caused by a) knockdown of DDR genes, b) genetic overexpression of upd3, c) genetic activation of JNK. But whether these effects are all related and reflect a linear pathway requires a little more work. For example, one prediction of the proposed model is that the increased susceptibility to oxidative stress-induced death in the hemocyte DDR gene knockdowns would be suppressed (perhaps partially) by simultaneous knockdown of upd3 and/or JNK. These types of epistasis experiments would strengthen the model and the paper.

As mentioned before, we had some technical difficulties combining the knockdown of bsk or upd3 with DDR genes. However, we added a new experiment in which we show that upd3null mutation can rescue the higher susceptibility of hemocytes with tefu and mei41 knockdown.

1. The (potential) connections between DDR/JNK/UPD3 and the oxidative stress effects on depletion of nutrient (lipids and glycogen) stores was also not fully developed. However, it may be the case that, in this paper, the authors just want to speculate that the effects of hemocyte DDR/upd3 manipulation on viability upon oxidative stress involve changes in nutrient stores.

In the revised version of the manuscript, we now provide a more thorough snRNA-seq analysis in the fat body upon PQ treatment to give more insights on the changes in the fat body upon PQ treatment. We added additional histological images of the abdominal fat body on control food and PQ food, to demonstrate the elimination of triglycerides from fat body with Oil-Red-O staining (Figure S1). We also analyzed now hemocyte-deficient (crq-Gal80ts>reaper) flies for their levels of triglycerides and carbohydrates during oxidative stress, to support our hypothesis that hemocytes are key players in the regulation of energy mobilization during oxidative stress. Loss of hemocytes (and therefore also their regulatory input on energy mobilization from the fat body) results in increased triglyceride storage in the fat body during steady state with a decreased consumption of these triglycerides on PQ food compared to control flies (Figure 1J). In contrast, glycogen storage and mobilization, which is mostly done in muscle, is not altered in these flies during oxidative stress (Figure 1L). Interestingly, free glucose levels are drastically reduced in hemocyte-deficient flies, which could be due to insufficient energy mobilization from the fat body and subsequently results in a higher susceptibility of these flies on oxidative stress (Figure 1K). Additionally, we aim to point out here that “functional” hemocytes are needed for effective response to oxidative stress, but this response has to be tightly balanced (see also new graphical abstract).

Reviewer #2 (Public Review):

Hersperger et al. investigated the importance of Drosophila immune cells, called hemocytes, in the response to oxidative stress in adult flies. They found that hemocytes are essential in this response, and using state-of-the-art single-cell transcriptomics, they identified expression changes at the level of individual hemocytes. This allowed them to cluster hemocytes into subgroups with different responses, which certainly represents very valuable work. One of the clusters appears to respond directly to oxidative stress and shows a very specific expression response that could be related to the observed systemic metabolic changes and energy mobilization. However, the association of these transcriptional changes in hemocytes with metabolic changes is not well established in this work. Using hemocyte-specific genetic manipulation, the authors convincingly show that the DNA damage response in hemocytes regulates JNK activity and subsequent expression of the JAK/STAT ligand Upd3. Silencing of the DNA damage response or excessive activation of JNK and Upd3 leads to increased susceptibility to oxidative stress. This nicely demonstrates the importance of tight control of JNK-Upd3 signaling in hemocytes during oxidative stress. However, it would have been nice to show here a link to systemic metabolic changes, as the authors conclude that it is tissue wasting caused by excessive Upd3 activation that leads to increased susceptibility, but metabolic changes were not analyzed in the manipulated flies.

We thank the referee for the suggestion to better connect upd3 cytokine levels to energy mobilization from the fat body. We agree that this is an important point to support our hypothesis. First, we added now a detailed analysis of fat body cells in our snRNA-seq data to evaluate the changes induced in the fat body upon oxidative stress. We further added additional metabolic analyses of hemocyte-deficient flies (crq-Gal80ts>reaper) to support our hypothesis that hemocytes are key players in the regulation of energy mobilization during oxidative stress (see also answer to referee 1). Loss of the regulatory role of hemocytes in the energy mobilization and redistribution leads to a decreased consumption of these triglycerides on PQ food compared to control flies (Figure 1J). In contrast, glycogen storage and mobilization from muscle, is not affected in hemocyte-deficient flies during oxidative stress (Figure 1L). Interestingly, free glucose levels are drastically reduced in hemocyte-deficient flies compared to controls, which could be due to insufficient energy mobilization from the fat body resulting in a higher susceptibility to oxidative stress (Figure 1K). This data supports our assumption that “functional” hemocytes are needed for effective response to oxidative stress, but this response has to be tightly balanced (see also new graphical summary).

The overall conclusion of this work, as presented by the authors, is that Upd3 expression in hemocytes under oxidative stress leads to tissue wasting, whereas in fact it has been shown that excessive hemocyte-specific Upd3 activation leads to increased susceptibility to oxidative stress (whether due to increased tissue wasting remains a question). The DNA damage response ensures tight control of JNK-Upd3, which is important. However, what role naturally occurring Upd3 expression plays in a single hemocyte cluster during oxidative stress has not been tested. What if the energy mobilization induced by this naturally occurring Upd3 expression during oxidative stress is actually beneficial, as the authors themselves state in the abstract - for potential tissue repair? It would have been useful to clarify in the manuscript that the observed pathological effects are due to overactivation of Upd3 (an important finding), but this does not necessarily mean that the observed expression of Upd3 in one cluster of hemocytes causes the pathology.

We agree with the referee that the pathological effects and increased susceptibility to oxidative stress are mediated by over-activated hemocytes and enhanced cytokine release, including upd3 during oxidative stress. We edited the revised manuscript accordingly to imply a “regulatory” role of upd3, which we suspect and suggest as an important mediator for inter-organ communication between hemocytes and fat body. Whereas our used model for oxidative stress (15mM Paraquat feeding) is a severe insult from which most of the flies will not recover, we could not account and test how upd3 might influence tissue repair after injury, insults and infection. We believe that this is an important factor, we aim to explore in future studies.

Reviewer #3 (Public Review):

In this study, Kierdorf and colleagues investigated the function of hemocytes in oxidative stress response and found that non-canonical DNA damage response (DDR) is critical for controlling JNK activity and the expression of cytokine unpaired3. Hemocyte-mediated expression of upd3 and JNK determines the susceptibility to oxidative stress and systemic energy metabolism required for animal survival, suggesting a new role for hemocytes in the direct mediation of stress response and animal survival.

Strength of the study:

1. This study demonstrates the role of hemocytes in oxidative stress response in adults and provides novel insights into hemocytes in systemic stress response and animal homeostasis.

1. The single-cell transcriptome profiling of adult hemocytes during Paraquat treatment, compared to controls, would be of broad interest to scientists in the field.

We are grateful to these positive comments on our data and are excited that the referee pointed out the importance of our provided snRNA-seq analysis of hemocytes and other cell types during oxidative stress. In the revised, version we now extended this analysis and looked not only into hemocytes but also highlighted induced changes in the fat body (Figure 4).

Weakness of the study:

1. The authors claim that the non-canonical DNA damage response mechanism in hemocytes controls the susceptibility of animals through JNK and upd3 expression. However, the link between DDR-JNK/upd3 in oxidative stress response is incomplete and some of the descriptions do not match their data.

In the revised manuscript, we aimed to strengthen the weaknesses pointed out by the referee. We now included additional genetic crosses to validate the connection of DDR signaling in hemocytes with upd3 release. For example, we added now survival studies where we show that upd3null mutation can rescue the higher susceptibility of flies with tefu and mei41 knockdown in hemocytes during oxidative stress. Furthermore, we added additional data to highlight the importance of hemocytes themselves as essential regulators of susceptibility to oxidative stress. We analyzed the hemocyte-deficient flies (crq-Gal80ts>reaper) for their triglyceride content and carbohydrate levels during oxidative stress (Figure 1 I-L). As outlined above, loss of hemocytes leads to a decreased consumption of these triglycerides on PQ food compared to control flies (Figure 1J). In contrast, glycogen storage and mobilization from muscle, is not affected in hemocyte-deficient flies during oxidative stress (Figure 1L). Interestingly, free glucose levels are drastically reduced in hemocyte-deficient flies, which could be due to insufficient energy mobilization from the fat body resulting in a higher susceptibility to oxidative stress (Figure 1K).

1. The schematic diagram does not accurately represent the authors' findings and requires further modifications.

We carefully revised the text throughout the manuscript describing our results and edited the graphical abstract to display that upd3 levels and hemocytes are essential to balance and modulate response to oxidative stress.

Reviewer #1 (Recommendations For The Authors):

The summary doesn't say too much about what the specific discoveries and results of the study are. The description is limited to just one sentence saying, "Here we describe the responses of hemocytes in adult Drosophila to oxidative stress and the essential role of non-canonical DNA damage repair activity in direct "responder" hemocytes to control JNK-mediated stress signaling, systemic levels of the cytokine upd3 and subsequently susceptibility to oxidative stress" which doesn't provide sufficient explanation of what the results were.

In the revised version of our manuscript, we now provide further information for the reader to outline the findings of our study in a concise way in the summary.

Reviewer #2 (Recommendations For The Authors):

1. To strengthen the conclusion that the DDR response suppresses JNK, and thus Upd3, rescue of DDR by upd3 null mutation would help (knockdown by Hml>upd3IR might not work, RNAi seems problematic).

We would like to thank the referee for this suggestion and included now a genetic experiment where we combined upd3null mutants with hemocyte-specific knockdown of mei-41 and tefu to test their susceptibility to oxidative stress. Our data indeed provide evidence that loss of upd3 rescues the higher susceptibility of flies with hemocyte-specific knockdown for tefu and mei-41 (Figure 6F). Furthermore, we see that upd3null mutants show a diminished susceptibility to oxidative stress compared to control flies (Figure 6F).

1. To link the observed effects to systemic metabolic changes, it would be useful to measure glycogen and triglycerides in these flies as well:

crq-Gal80ts>reaper to see what role hemocytes play in the observed metabolic changes.

Hml-Upd3 overexpression and Upd3 null mutant (Upd3 RNAi seems to be problematic, we have similar experiences) to see if Upd3 overexpression leads to even more profound changes as suggested, and if Upd3 mutation at least partially suppresses the observed changes.

We agree with the referee that analyzing the connection of hemocyte activation to metabolic changes should be demonstrated in our manuscript to support our claim that hemocytes are important regulators of energy mobilization during oxidative stress. Hence, we analyzed triglycerides and carbohydrate levels in hemocyte-deficient flies (crq-Gal80ts>reaper) during oxidative stress. Indeed, we found substantial differences in energy mobilization in these flies supporting the assumption that the higher susceptibility of hemocyte-deficient flies could be caused by substantial decrease in free glucose and inefficient lysis of triglycerides from the fat body (Figure 1I-K).

1. To test whether the cause of the increased susceptibility to oxidative stress is due to Upd3 overactivation induced by DDR silencing, the authors should attempt to rescue DDR silencing with an Upd3 null mutation.

The suggestion of the reviewer was included in the revised manuscript and as outlined above we now added this data set to our manuscript (Figure 6F). Indeed, we can now provide evidence that upd3null mutation rescues the higher susceptibility of flies with DDR knockdown in hemocytes.

1. Lethality after PQ treatment varies widely (sometimes from 10 to 90%! as in Figure 5D) - is this normal? In some experiments the variability was much lower. In particular, Figure 5D is very problematic and for example the result with upd3 null mutant compared to control is not very convincing. This could be an important result to test whether Upd3, with normal expression likely coming from cluster 6, actually plays a beneficial role, whereas overexpression with Hml leads to pathology.

We agree with the referee that it would be more convincing if the variation cross of survival experiments would be less. However, we included a lot of flies and vials in many individual experiments to test our hypothesis and variation in these survivals was always the case. These effects can be caused by many factors for example the amount of food intake by the flies, genetic background or inserted transgenes. The n-number is quite high across our survivals; so that we are convinced, the seen effects are valid. This reflects also the power of using Drosophila melanogaster as a model organism for such survivals. The high n-number in our data falls into a normal Gauss distribution with a distinct mean susceptibility between the genotypes analyzed.

1. I like the conclusion at the end of the results: line 413: "We show that this oxidative stressmediated immune activation seems to be controlled by non-canonical DNA damage signaling resulting in JNK activation and subsequent upd3 expression, which can render the adult fly more susceptible to oxidative stress when it is over-activated." This is actually a more appropriate conclusion, but in the summary, introduction and discussion along with the overall schematic illustration, this is not actually stated as such, but rather as Upd3 released from cluster 6 causes the pathology. For example: line 435 "Hence, we postulate that hemocyte-derived upd3, most likely released by the activated plasmatocyte cluster C6 during oxidative stress in vivo and subsequently controlling energy mobilization and subsequent tissue wasting upon oxidative stress."

We thank the referee for this suggestion and edited our manuscript and conclusions accordingly.

Reviewer #3 (Recommendations For The Authors):

1. In Figure 2, the authors claim showed that PQ treatment changes the hemocyte clusters in a way that suppresses the conventional Hml+ or Pxn+ hemocytes (cluster1) while expanding hemocyte clusters enriched with metabolic genes such as Lpin, bmm etc. It is not clear whether these cells are comparable to the fat body and if these clusters express any of previously known hemocyte marker genes to claim that these are bona fide hemocytes.

We now included a new analysis of our snRNA-seq data in Figure S4, where we clearly show that all identified hemocyte clusters do not have a fat body signature and are hemocytes, which seem to undergo metabolic adaptations (Figure S4A). Furthermore, we show that the identified fat body cells have a clear fat body signature (Figure S4B) and do not express specific hemocyte markers (Figure S4C).

1. In Figure 4C, the authors showed that comet assays of isolated hemocytes result in a statistically significant increase in DNA damage in DDR-deficient flies before and after PQ treatment. However, the authors conclude that, in lines 324-328, the higher susceptibility of DDR-deficient flies is not due to an increase in DNA damage. To explicitly conclude that "non-canonical" DNA damage response, without any DNA damage, is specifically upregulated during PQ treatment, the authors require further support to exclude the potential activation of canonical DDR.

The referee is correct that we do not provide direct evidence for non-canonical DNA damage signaling. Therefore, we also decided to tune down our statement here a bit and removed that claim from the title. Increase in DNA damage can of course also increase the non-canonical DNA damage signaling pathway, loss of DNA damage signaling genes such as tefu and mei-41 seem to only have minor impacts on the overall amount of DNA damage acquired in hemocytes by oxidative stress. We therefore concluded that the induction in immune activation is most unlikely only caused by increased DNA damage but might be connected to dysregulation in non-canonical DNA damage signaling. Canonical DNA damage signaling leads essentially to DDR, which could be slow in adult hemocytes because they post-mitotic, or to apoptosis, which we could not observe in the analyzed time window in our experiments. Hemocyte number remained stable over the 24h PQ treatment without reduction in cell number (Figure 1H).

1. From Figure 4D-F, the authors showed that loss of DDR in hemocytes induces the expression of unpaired 2 and 3, Socs36E, which represent the JAK/STAT pathway, and thor, InR, Pepck in the InR pathway, and a JNK readout, puc. These results indicate that the DDR pathway normally inhibits the upd-mediated JAK/STAT activation upon PQ treatment, compared to wild-type animals during PQ treatment in Figure 1B-C, which in turn protects the animal during oxidative stress responses. However, the authors claim that "enhanced DNA damage boosts immune activation and therefore susceptibility to oxidative stress (lines 365-366); we show that this oxidative stress-mediated immune activation seems to be controlled by non-canonical DNA damage signaling resulting in JNK activation and subsequent upd3 expression (line 413-416)". These conclusions are not compatible with the authors' data and may require additional data to support or can be modified.

In the revised manuscript, we carefully revised now the text and our statements that it seems that DNA damage signaling in hemocytes has regulatory or modulatory effect on the immune response during oxidative stress. Accordingly, we also adjusted our graphical summary. We agree with the referee and used the term “non-canonical” DNA damage signaling more carefully throughout the manuscript. The slight increase in DNA damage seen after PQ treatment can contribute to immune activation but seems to be not correlative to the induced cytokine levels or the susceptibility of the flies to oxidative stress.

1. In Fig 1I, the authors showed that genetic ablation of hemocytes using UAS-repear induces susceptibility to PQ treatment. It is possible that inducing cell death in hemocytes itself causes the expression of cytokine upd3 or activates the JNK pathway to enhance the basal level of upd3/JNK even without PQ treatment. If this phenotype is solely mediated by the loss of hemocytes, the results should be repeated by reducing the number of hemocytes with alternative genetic backgrounds.

In the different genotypes analyzed across our manuscript we did not detect cell death of hemocytes or a dramatic reduction in hemocytes number (see Figure 1H, Figure 5B, Figure 6C). The higher susceptibility if hemocyte-deficient flies during oxidative stress is most likely caused by the loss of their regulatory role during energy mobilization. We tested triglyceride levels in hemocyte-deficient flies and found a decreased triglyceride consumption (lipolysis), with reduced levels of circulating glucose levels. This findings support our hypothesis that hemocytes are needed to balance the response to oxidative stress. In contrast, the flies with DDR-deficient hemocytes show higher systemic cytokine levels, which most likely enhance energy mobilization from the fat body and therefore result in a higher susceptibility of the fly to oxidative stress. Hence, we claim that hemocytes and their regulation of systemic cytokine levels are important to balance the response to oxidative stress and guarantee the survival of the organism.

1. Lethality of control animals in PQ treatment is variable and it is hard to estimate the effect of animal susceptibility during 15mM PQ feeding. For example, Fig1A shows that control animals exhibit ~10% death during 15mM PQ which is further enhanced by crq-Gal80>reaper expression to 40% (Fig 1I). However, in Fig 5D-E, the basal lethality of wild-type controls already reaches 40~50%, which makes them hard to compare with other genetic manipulations. Related to this, the authors demonstrated that the expression of upd3 in hemocytes is sufficient to aggravate animal survival upon PQ treatment; however, upd3 null mutants do not rescue the lethality, which indicates that upd3 is not required for hampering animal mortality. These data need to be revisited and analyzed.

As outlined above, we find the variability of susceptibility to oxidative stress across all of our experiments. This could be due to different effects such as food intake but also transgene insertion and genetic background. Crq-gal80ts>reaper flies are healthy, but show a shortened life span on normal food (Kierdorf et al., 2020) due to enhanced loss of proteostasis in muscles. We show in the revised manuscript that these flies have a higher susceptibility to oxidative stress and that this effect could be mediated by defects in energy mobilization and redistribution as shown by less triglyceride lysis from the fat body and decreasing levels in free glucose. This would explain the high mortality rate of these flies at 7 days after eclosion. Paraquat treatment (15mM) is a severe inducer of oxidative stress, which results in death of most flies when they are maintained for longer time windows on PQ food. Hence, it is a model, which is not suitable to examine and monitor recovery from this detrimental insult. upd3null mutants were extensively reexamined in this manuscript, and even though we could not see a full protection of these flies from oxidative stress induced death, we found a reduced susceptibility compared to control flies (Figure 6F). Furthermore, when we combined upd3null mutants with flies deficient for tefu and mei-41 in hemocytes, the increased susceptibility to oxidative stress was rescued.

Version 2

Abstract

Summary

Environmental factors, infection, or injury can cause oxidative stress in diverse tissues and loss of tissue homeostasis. Effective stress response cascades, conserved from invertebrates to mammals, ensure reestablishment of homeostasis and tissue repair. Hemocytes, the Drosophila blood-like cells, rapidly respond to oxidative stress by immune activation. However, the precise signals how they sense oxidative stress and integrate these signals to modulate and balance the response to oxidative stress in the adult fly are ill-defined. Furthermore, hemocyte diversification was not explored yet on oxidative stress. Here, we employed high throughput single nuclei RNA-sequencing to explore hemocytes and other cell types, such as fat body, during oxidative stress in the adult fly. We identified distinct cellular responder states in plasmatocytes, the Drosophila macrophages, associated with immune response and metabolic activation upon oxidative stress. We further define oxidative stress-induced DNA damage signaling as a key sensor and a rate-limiting step in immune-activated plasmatocytes controlling JNK-mediated release of the pro-inflammatory cytokine unpaired-3. We subsequently tested the role of this specific immune activated cell stage during oxidative stress and found that inhibition of DNA damage signaling in plasmatocytes, as well as JNK or upd3 overactivation, result in a higher susceptibility to oxidative stress. Our findings uncover that a balanced composition and response of hemocyte subclusters is essential for the survival of adult Drosophila on oxidative stress by regulating systemic cytokine levels and cross-talk to other organs, such as the fat body, to control energy mobilization.

Introduction

Metazoan organisms must constantly adapt to the surrounding environment. Environmental factors, including radiation, pathogens or toxins, permanently lead to the generation of reactive oxygen species (ROS)1. In vertebrates and invertebrates, low levels of ROS support cellular pathways like immune cell differentiation and even immune function including antimicrobial defense2,3. At high concentrations, ROS lead to oxidative stress, a deleterious process that negatively impacts protein synthesis and stability, lipid metabolism as well as overall genome stability. ROS mediated cellular damage can further result in degenerative diseases or cancer4. Therefore, metazoan organisms employ efficient repair and defense mechanisms to circumvent high levels of oxidative stress and subsequent organ damage. Hence, keeping the balance between the production and the antioxidant-dependent clearance of ROS is essential for the survival and well-being of the affected organism5. Oxidative stress induced tissue damage is most often followed by a strong local immune response. These molecular processes are conserved across vertebrates and invertebrates such as the fruit fly Drosophila melanogaster6.

Hemocytes, the Drosophila blood cells, are essential cellular components of the immune and stress response machinery. They consist of three different subtypes: crystal cells, lamellocytes and plasmatocytes7. Plasmatocytes are considered to be macrophage equivalents in Drosophila and comprise up to 95 % of all hemocytes in the adult fly8. They share many functions as well as the multilayered ontogeny with vertebrate tissue macrophages9,10. Adult plasmatocytes are actively involved in many immune functions such as pathogen recognition, defense and phagocytosis but also chronic immune activation upon metabolic stress by dietary lipids11,12. Recent studies highlighted the essential role of adult plasmatocytes as key players in regulation of tissue homeostasis, including the gut by regulating the intestinal stem cell niche or in muscles by controlling glucose homeostasis13,14.

Similar to higher eukaryotes, levels of ROS are used as a central antimicrobial defense mechanism of hemocytes against invading pathogens15. Moreover, larval hemocytes respond differentially to ROS, with two cell subsets mounting a biphasic immune reaction to invading bacteria by controlling hemocyte spreading and adhesion as well as crystal cell rupture15. Larval hemocytes have been also attributed to respond to increased oxidative stress of other tissues, as seen in response to oxidative stress in the eye-antenna imaginal disc16. Similarly, by regulating the intestinal stem cell (ISC) proliferation in the damaged midgut during infection-induced or chemical-induced oxidative stress14, adult hemocytes were described to release the bone morphogenetic protein (BMP) homologue decapentaplegic (dpp) to facilitate tissue repair14. Other studies pointed to an essential role of hemocyte-derived unpaired-3 (upd3), which is the Drosophila orthologue of Interleukin-6, in the regulation of ISC proliferation during septic injury17, as well as wound healing during sterile injury18. In line with these findings, hemocytes have been implicated in organ-to-organ immune signaling for example upon infection-induced injury in the gut and antimicrobial peptide (AMP) production in the fat body19. Hemocyte activation is thus a hallmark of oxidative stress in defined tissues, though their role in modulating and controlling oxidative stress response to facilitate tissue repair and survival of the organism remains undefined.

Here, we demonstrate an essential role of hemocytes to control susceptibility to oxidative stress, where hemocytes orchestrate systemic oxidative stress response by defined cellular states identified by unbiased transcriptomic profiling. Plasmatocytes segregate in oxidative stress responder subclusters, characterized by immune activation and cellular stress response, and potential bystander subsets associated with metabolic gene regulations. Mechanistically, we unravel that the direct immune response to oxidative stress by plasmatocytes is driven by c-Jun N-terminal kinase (JNK) and Jak/STAT pathway activation with subsequent induction of the pro-inflammatory cytokine upd3. This inflammatory cascade is to a certain amount regulated by DNA damage signaling in plasmatocytes and essentially defines the susceptibility of the fly to oxidative stress. Deficiencies in the DNA damage signaling machinery in plasmatocytes lead to elevated JNK activation, cytokine release and subsequently a higher susceptibility of the adult fly to oxidative stress. Hence, our data point to a key role of plasmatocytes in balancing systemic cytokine levels and stress response upon oxidative stress. Collectively we provide new insights on the role of macrophages in orchestrating systemic stress responses in other tissues and to balance tissue wasting and energy mobilization.

Results

Hemocytes control susceptibility to oxidative stress

In vertebrates and invertebrates, prolonged oxidative stress can cause severe tissue damage, hence controlled activation of stress signaling cascades are essential for the survival of the organism and the resolution of oxidative stress with subsequent tissue repair. Yet the precise role of macrophages in the initiation and control of oxidative stress signaling needs to be determined.

To analyze the role of macrophages in signaling cascades to oxidative stress, we took advantage of the well-established Paraquat (PQ)-feeding model in adult Drosophila melanogaster to induce systemic oxidative stress20. Feeding of 7-days old w;Hml-Gal4,UAS-2xeGFP/+ (Hml/+) flies with PQ-supplemented sucrose solution led to an increased lethality of adult Drosophila after 18 hours in a dose dependent manner (0mM: 1.25 ± 1.25 %, 2mM: 2.5 ± 1.6 4%, 15mM: 27.78 ± 5.47 %, and 30mM: 38.75 ± 6.67 %) (Figure S1A). In agreement with previous studies20, 15mM PQ was chosen for further experiments (Figure 1A). First, we determined which systemic changes are associated with susceptibility to oxidative stress. Infection induced damage in enterocytes can result in gut leakage and subsequent death17. We tested if the death of flies after 18 hours on 15mM PQ food was associated with PQ-induced damage of gut enterocytes and resulting gut leakage, but we found no indication of gut leakage on 15mM PQ food (Figure S1B). Only very few flies on control food and PQ indicated gut leakage of the food color to thorax and abdomen which was unrelated to the treatment and not altered between the groups (control: 3.75%, 2mM PQ: 2.53 %, 15mM PQ: 1.11 % and 30mM PQ: 3.75 %) (Figure S1B). We investigated if aberrant systemic induction of immune signaling cascades are seen upon PQ-mediated oxidative stress. For this, gene expression levels of different known immune activated genes were analyzed by qPCR such as unpaired (upd) cytokines (upd1, upd2, upd3), Jak/STAT down-stream targets (Turandot-A (TotA), suppressor of cytokine signaling at 36E (Socs36E)), stress induced JNK activation (puckered (puc)) (Figure 1B), as well as expression of nuclear factor ‘kappa-light-chain-enhancer’ of activated B-cells (Nfkb)-induced antimicrobial peptides (AMP) and further cytokines such as eiger, dpp and dawdle (daw) (Figure S1C-D). These analyses did not reveal significant differences for these gene groups on a systemic level. In contrast, genes associated with carbohydrate metabolism such as insulin receptor (InR), phosphoenolpyruvate carboxykinase (Pepck), and Thor were significantly upregulated upon PQ-feeding (Figure 1C). These findings point to an activation of forkhead box, sub-group O (foxo)-target genes and a potentially reduced insulin signaling, despite the lack of significant alterations in the expression of the Insulin-like peptides (Ilps) Ilp2, Ilp3 and Ilp5 (Figure 1C).

Loss of hemocytes increases the susceptibility of adult Drosophila to oxidative stress by Paraquat feeding.

(A) Survival of Hml/+ flies with control food (C) (n=20) and 15mM Paraquat (P) (n=28) at 29°C. Five independent experiments were performed. Student’s unpaired t-test: ****p<0.0001, each data point represents a sample with 10 flies. (B-C) RT-qPCR of whole flies were performed to investigate gene-expressions of the JAK-STAT-signaling pathway (B) and insulin-signaling pathway (C). All transcript levels were normalized to the expression of the loading control Rpl1 and are shown in arbitrary units [au]. Each data point (n=5) represents a sample of three whole flies. Student’s unpaired t-tests were performed for each transcript to compare the gene-expression between control and treated flies: **p<0.01; ***p<0.001; ****p<0.0001. (D) Glucose and (E) glycogen levels of whole flies were measured in control and PQ-treated flies. Each dot (n=3) represents a sample containing three flies. Student’s unpaired t-test: **p<0.01. (F) Triglyceride (TG) amounts were determined via thin-layer chromatography (TLC). Left panel: Representative image of a TLC plate, each band represents a sample with ten flies. Right panel: Quantification thereof. n= 3 per group were analyzed. Student’s unpaired t-test: **p<0.01. (G) Representative images of 7 days old HmlΔ4>eGFP flies treated with 5% sucrose solution and 15mM PQ respectively. Scale bars = 500µm. (H) Hemocyte quantification of 7 days old HmlΔ4>eGFP flies treated with control food (5% sucrose solution) (n=5) or 15mM Paraquat (n=5). Each data point represents one fly. Statistical significance was tested using student’s unpaired t-test: p=0.508. (I) Survival of crq-Gal80ts/+ control flies (n=13) and hemocyte depleted crq-Gal80ts>reaper flies (n=7) on control food (n=20) and 15mM Paraquat (n=28) at 29°C. Student’s unpaired t-test: ****p<0.0001, each data point represents a sample with 10 flies. (J) TG amounts in crq-Gal80ts/+ control flies and hemocyte-deficient crq-Gal80ts>reaper flies on control food or Paraquat food. Left panel: Representative image of a TLC plate, each band represents a sample with ten flies. Right panel: Quantification thereof. n= 3 per group were analyzed. One-way ANOVA: *p<0.05. (K) Glucose and (L) Glycogen levels of whole flies were measured in crq-Gal80ts/+ control flies and hemocyte-deficient crq-Gal80ts>reaper flies. Each dot (n= 3) represents a sample with three whole flies. One-way ANOVA: ****p<0.0001.

To test the consequences for energy mobilization and nutrient storage, we measured free glucose content, as well as glycogen and triglyceride levels in whole flies after 18 hrs on control or 15mM PQ. Although no changes in free glucose levels (control: 1.49±0.02µg and PQ: 1.50±0.07µg, ns p=0.954) (Figure 1D) were found, we observed a clear decrease in stored glycogen (control: 1.02±0.07µg and PQ: 0.33±0.08µg, **p=0.003) (Figure 1E) and stored triglycerides (Figure 1F) in PQ-treated flies compared to control. Next, we examined histological sections of flies to define which tissues might be affected. Hematoxylin and eosin (H&E) and Oil Red O (ORO) stainings showed a shrinkage in fat body cell size and stored lipid-filled vesicles in the abdominal fat body compared to flies kept under control conditions (Figure S1E-F). Fat body development, energy mobilization and storage have been recently connected to the presence, cell number and subsequent cross-talk of hemocytes with fat body21, therefore we quantified GFP+ hemocytes in Hml/+ flies after 18 hrs on PQ or control food (Figure 1G-H). Our analysis did not reveal alterations in hemocyte numbers and especially no oxidative stress induced loss of hemocytes (control: 243.4±24.33 hemocytes per fly and PQ: 219.6±24.22 hemocytes per fly, ns p=0.508) (Figure 1G-H). To evaluate if hemocytes are essential contributors to the stress-mediated metabolic changes and susceptibility to oxidative stress, we used “hemocyte-deficient” crq-Gal80ts>reaper (w;tub-Gal80ts/UAS-CD8-mCherry;crq-Gal4/UAS-reaper) flies and tested their survival rate compared to control flies on PQ food (Figure 1I). Hemocyte-deficient flies displayed a significantly higher mortality rate on 15mM PQ (43.33±6.30%) compared to control flies (13.08±2.63%) (Figure 1I). Next, we analyzed triglycerides, as well as free glucose and glycogen levels in whole flies after 18 hrs on control or 15mM PQ. Interestingly, we found that hemocyte-deficient flies compared to the control genotype mobilized less triglycerides from the fat body during oxidative stress (Figure 1J). When we looked into free glucose levels, we found that these flies show a decrease in free glucose upon PQ feeding, whereas the control genotype showed similar free glucose levels as on normal food (Figure 1K). In contrast we did not detect any difference in glycogen levels between the two genotypes on control food and on PQ food (Figure 1L). Our findings show detrimental changes in energy mobilization upon oxidative stress in the absence of profound systemic overactivation of immune signaling cascades, however the presence or absence of hemocytes seem to be a key determinant of the susceptibility to oxidative stress and results in altered energy mobilization in adult Drosophila. This strongly points to an essential role for hemocytes in controlling stress response and organismal survival upon oxidative stress.

Paraquat exposure generates diverse transcriptional states of adult hemocytes with a specific cell state associated with oxidative stress response

To gain further insights into the alterations induced in the adult fly upon oxidative stress and especially to the hemocyte compartment, we decided to take advantage of unbiased single nucleus RNA-sequencing (snRNA-seq). We isolated nuclei from control food and PQ-treated Hml-DsRed.nuc flies with a modified Frankenstein protocol22 and performed fluorescence activated nuclear sorting of DAPI+Draq7+ nuclei (Figure S2A). From two independent sorts we obtained a total of 29000 nuclei which were utilized for subsequent droplet-based barcoding of single nuclei, and generation of single nucleus cDNA libraries. We could successfully obtain gene expression data from 11839 nuclei (Figure S2B-D). Quality control analysis followed by unsupervised hierarchical clustering of all analyzed nuclei using Seurat package23 revealed 20 distinct nuclei clusters (C) with differential gene expression (Figure 2A). Cell identity was annotated according to the cell type-specific signatures determined by the Fly Cell Atlas consortium (Figure 2A)24. All annotated cell types were detected in PQ-treated and control flies (Figure 2B).

Unbiased single nuclei transcriptomic profiling identified diverse transcriptomic states of hemocytes associated with oxidative stress response.

(A) Uniform Manifold Approximation and Projection (UMAP) visualization of single nucleus states from flies exposed to 15mM PQ containing or control food. Dashed lines indicate the broad cell types. Male accessory gland main cells (MAG), Malpighian tubule principal cells (MTC), male germline cells (MGC), outer photoreceptor cells (PRC), tracheolar cells (TC), epithelial cells (EC). Colors indicate distinct clusters. n=11839 nuclei are shown. (B) UMAP visualization of nuclei from A colored by treatment. Nuclei from control group are labeled in red and nuclei from PQ group are labeled in blue. Dashed lines indicate the broad cell type as A. (C) UMAP visualization of hemocytes from A after sub-setting and re-clustering. Colors and dashed lines indicate distinct clusters. n=1354 nuclei are shown. (D) UMAP visualization of hemocyte clusters colored by treatment. Nuclei from control group are labeled in red and nuclei from PQ group are labeled in blue. Dashed lines indicate distinct clusters. n=1354 nuclei are shown. (E) Heat map of top 20 DE genes for each hemocyte cluster. Gene names are indicated on the left. Colors in the heat map correspond to normalized scaled expression. (F) Volcano plots comparing pseudo bulk gene expression of individual hemocyte cluster vs. all hemocytes. The –log10-transformed adjusted P value (P adjusted, y-axis) is plotted against the log2-transformed fold change (FC) in expression between the indicated cluster vs remaining hemocytes (x-axis). Non-regulated genes are shown in grey, not significantly regulated genes are shown in green, significantly regulated genes with a FC<1.5 are shown in blue and significantly regulated genes with a FC>1.5 are shown in red.

Using cell type-specific signatures, three of the clusters were annotated as hemocytes (clusters: 2, 11 and 14) (Figure 2A-B). Re-scaling and clustering of all 1354 identified hemocyte nuclei revealed 8 distinct clusters (Figure 2C). We verified hemocyte-specific signature genes across these clusters including genes such as the transcription factor Serpent (Srp) and one or more of the following hemocyte signature genes: Hemolectin (Hml), croquemort (crq), Nimrod C1 (NimC1), eater, and Peroxidasin (Pxn) (Figure S3A-L)25,26. Cluster 8 expressed various crystal cell-specific markers including lozenge (lz), pebbled (peb), prophenoloxidase 1 (PPO1), and prophenoloxidase 2 (PPO2), whereas all other clusters seemed to be plasmatocytes (Figure S3A-L). In line with previous reports, the embryonic and larval hemocyte markers Hemese (He) (data not shown) and singed (sn) were not expressed across any of the identified clusters (Figure S3H). Interestingly, when we defined which hemocyte clusters were enriched in control flies or PQ-treated flies, the two plasmatocyte clusters C1 and C7 were primarily derived from control flies, whereas C2, C3, C4 and C6 are predominantly found in PQ-treated flies (Figure 2D and Figure S3M). The plasmatocyte cluster C5 as well as the crystal cell cluster C8 presented in both conditions to similar proportions (Figure 2D and Figure S3M).

Next, we analyzed the TOP20 differentially expressed (DE) genes across all hemocyte clusters. We found clear differences between the 7 distinct plasmatocyte clusters and the crystal cell cluster (Figure 2E and Table S1). The crystal cell cluster C8 showed a high expression of crystal cell specific genes including lz, peb, PPO1 and PPO2 (Figure 2E). Plasmatocyte cluster C1, which was mostly present in control conditions, had a remarkable expression of hemocyte specific marker genes such as Hml and Pxn, as well as migration-associated genes including the semaphorins Sema2a and Sema5c (Figure 2E). Yet this cluster also showed higher expression of genes contributing to metabolic regulations such as the L-gulonate 3-dehydrogenase Had2 and the gluconolactonase regucalcin (Figure 2E). The cluster C7 and C1, mostly found in control flies, substantially shared gene expression profiles. In contrast, cluster C2 and C3 were nearly exclusively found in PQ-treated flies and showed expression of genes involved in metabolic adaptations. Due to this finding we verified that these cell clusters are bona fide hemocytes and not fat body cells. We analyzed the expression of characteristic fat body signature genes, such as Trehalose-6-phosphate synthase 1 (Tps1), CG16758, CG2233, CG4716, Ultrabithorax (Ubx), Calcium/calmodulin-dependent protein kinase I (CaMKI), Desaturase 1 (Desat1), Phosphoribosyl pyrophosphate synthetase (Prps), CG34166 and Apolipoprotein lipid transfer particle (Apoltp) across all identified hemocyte clusters (Figure S4A). Metabolic adaptation within these clusters included high expression of the glucose transporter 4 enhancer factor (Glut4EF) or the phosphatidate phosphatase Lipin (Lpin) in C2 or the induced expression of the lipase brummer (bmm), the autophagy-associated gene Atg17 or the lipophorin receptor LpR1, involved in uptake of neutral lipids from the hemolymph, in C3 (Figure 2E). Overall gene expression profiles in clusters C2 and C3 showed an induction of insulin signaling, increased autophagy and lipolysis pointing to adaptation of energy mobilization in these clusters upon PQ treatment (Figure 2E). Cluster C4 was characterized by a remarkable upregulation of genes involved in translation including many ribosomal proteins such as RpL8, RpL31 or RpS13 (Figure 2E). Furthermore, C4 included genes associated with unsaturated fatty acid metabolism such as Desaturase 1 (Desat1). Similarly, cluster C5 also showed induction of genes involved in fatty acid metabolism such as ELOVL fatty acid elongase 7 (ELOVL) (Figure 2E). Of note, we also found Tie-like receptor tyrosine kinase (Tie) specifically expressed in this cluster, which is described to bind PDGF– and VEGF-related factor (Pvf) ligands and contribute to cell survival and migration, as well as actin-associated binding proteins including polychaetoid (pyd) and formin 3 (form3). All these gene expression changes are pointing to a plasmatocyte cluster with a gene expression profile associated with mobility and migration. Among all plasmatocyte clusters, we found cluster C6 which presented a strong immune activation phenotype with induction of Jak/STAT target genes including Fasciclin 3 (Fas3) and Turandot-M (TotM) among its TOP20 DE genes, but also activation of oxidative stress response genes such as the heat shock proteins Hsp23, Hsp26 and Hsp27 or cytoskeletal genes including myosin Myo31DF or spire (spir) (Figure 2E).

Due to the partially shared expression profile of genes across hemocyte clusters, we wondered if there were genes exclusively expressed in individual clusters. However, cluster C1 showed that Sema2a seems to be exceptionally high expressed in these homeostatic plasmatocytes and genes associated with metabolic regulation (e.g. InR, Lpin or bmm) or Jak/STAT activation (e.g. Fas3), as seen in PQ-associated clusters, are downregulated (Figure 2F). Cluster C5 showed enrichment for genes such as the transcription factor grainy head (grh) but also cytoskeletal-associated genes such as jitterbug (jbug) and form3, suggesting a migratory potential of plasmatocytes from this cluster within their host tissues (Figure 2F). Interestingly, plasmatocyte cluster C6 presented with the most profound gene expression changes compared to all other clusters. Similar to our previous observations, we identified that genes specifically expressed in this cluster were associated with Jak/STAT pathway activation (e.g. Fas3, TotM), cytoskeletal rearrangements and migration (e.g. Myo31DF, spir or cheerio (cher)) and response to oxidative stress (e.g. Nicotinamide amidase (Naam) or Hsp23) (Figure 2F). We also included the crystal cell cluster C8 in this analysis and verified unique expression of crystal cell markers compared to the seven plasmatocyte clusters (Figure 2F). Our snRNA-seq analysis revealed that adult hemocytes respond differentially to oxidative stress with a specific cluster of plasmatocytes associated with immune activation and stress response. Despite our findings that on systemic level we did not detect a strong induction of immune activation upon oxidative stress, we now provide evidence that plasmatocyte cluster C6 showed a substantial immune response towards oxidative stress.

Direct response to oxidative stress in vitro mimics the transcriptional changes seen in cluster C6 plasmatocytes in vivo

Plasmatocytes responded to PQ-mediated oxidative stress in different cellular states. We defined cluster C6 as a cell state which is associated with oxidative stress response and immune activation. Since cluster C6 showed gene expression associated with oxidative stress and immune activation we focused our further analysis on this cluster. To further substantiate our findings and address the signaling pathways that could lead to diverse transcriptional states, we inferred the transcription factor (TF) network based on gene expression data (Figure 3A). This analysis revealed that genes with binding sites for Xbp1 were enriched. Xbp1 is a transcription factor associated with stress response. In addition to this, the Nf-κB-related transcription factor Relish (Rel) and kayak (kay), a component of the activator protein-1 (AP-1) transcription factor complex induced by JNK signaling, were also enriched in this cluster (Figure 3A). To determine if the gene expression signature of cluster C6 plasmatocytes reflects a functional state of plasmatocytes directly induced by oxidative stress, we treated S2 cells, a Drosophila plasmatocyte cell line, with different concentrations of PQ (15mM and 30mM) in vitro (Figure S5). First, we performed flow cytometric analysis of ROS levels via CellROX staining after 6 and 24 hours PQ treatment compared to controls (Figure S5A-B). We observed a dose-dependent increase in intracellular ROS at all time points analyzed with higher ROS levels at 6 hrs (0mM: 1903±64 median fluorescence intensity (MFI), 15mM: 6735±1096 MFI, and 30mM: 7435±237 MFI) compared to 24 hrs (0mM: 1236±135 MFI, 15mM: 4456±228 MFI, and 30mM: 5557±303 MFI) (Figure S5C). Increase in ROS and oxidative stress can lead to cellular damage such as damage in nuclear and mitochondrial DNA.

A specific cluster of plasmatocytes responds to oxidative stress with immune activation by Jak/STAT, DDR and JNK signaling.

(A) Key transcription factors (TFs) regulating hemocytes states. Heat map of scaled TF activity at a single cell level in hemocytes states from Figure 2C. Colors in the heat map correspond to scaled expression. Numbers on top indicate the cluster identity. Labels on the left and right indicate the TFs. (B) Feature plots showing scaled expression of selected genes associated with JAK/STAT signaling (upd3, Socs36E, TotA, TotM, Fas3), JNK and DNA damage signaling (puc, Gadd45), insulin signaling (Thor, InR), TGFβ signaling (dpp, daw) and TNF signaling (egr).

We next tested if plasmatocytes directly exposed to oxidative stress might also experience this type of stress response by performing comet assays of treated and untreated S2 cells. Upon PQ treatment, S2 cells showed a remarkable and dose-dependent increase in DNA damage compared to untreated cells (0mM: 101.0±2.1, 5mM: 117.6±3.3, 10mM: 136.6±3.8, and 15mM: 147.8±5.5, data shown as olive tail moment) (Figure S5D). We were then wondering if PQ-treated S2 cells would resemble cluster C6 plasmatocytes in vivo and performed gene expression analysis of S2 cells treated with 15mM PQ after 6 and 24 hrs. Comparable to our results for cluster C6 in vivo, we found a profound induction of the Jak/STAT target genes by qRT-PCR. In S2 cells, we also detected a significant increase of the Jak/STAT cytokines upd1, upd2 and upd3 (Figure S5E), which we did not see before in vivo. Furthermore, we detected an increased expression of the JNK-target gene puckered (puc), but barely any induction of metabolic genes (Figure S5E). All these results further support our hypothesis that direct exposure of S2 cells with PQ in vitro mimics gene expression changes seen in a particular activated cell stage of plasmatocytes upon oxidative stress in vivo.

Because we found enrichment of genes associated with the Nf-κB-related transcription factor Rel in the immune-activated plasmatocyte cluster C6, we further verified if any other cytokines or AMPs are induced in PQ-treated S2 cells. We identified a slight but significant induction of the cytokine daw, but no other alterations in dpp or the AMPs Drs, Dro or Mtk (Figure S5F). The profound induction of Jak/STAT target genes together with the activation of JNK target genes and oxidative stress induced DNA damage in S2 cells in vitro highly support our hypothesis that cluster C6 plasmatocytes reflect a cellular state associated with the direct response to oxidative stress in vivo. Along this line, we went back to our snRNA-seq data and specifically searched for Jak/STAT target genes (upd3, Socs36E, TotA, TotM and Fas3) across all clusters (Figure 3B). Here we identified an enriched expression of these genes in cluster C6, and a unique expression of the pro-inflammatory cytokine upd3 in cluster C6 which was not detected in any other hemocyte cluster (Figure 3B). Of note, we did not detect elevated upd3 levels in gene expression profiling of whole flies (Figure 1B), but now unraveled elevated upd3 induction in the particular hemocyte cell cluster C6 upon PQ treatment. Interestingly, we also identified that Growth arrest and DNA damage-inducible 45 (Gadd45), a gene linking DNA damage signaling and stress induced JNK signaling, was enriched in plasmatocyte cluster C6 (Figure 3B). This could indicate that plasmatocytes within cluster C6 undergo oxidative stress mediated DNA damage, which results in activation of the DNA damage repair machinery but also JNK activation, similar to what we have seen in S2 cells in vitro. However within our snRNA-seq analysis, the JNK target gene puc was expressed across all hemocyte clusters analyzed in vivo without a specific induction in cluster C6 (Figure 3B). We looked again into the expression of metabolic genes including Thor and InR as well as other cytokines such as dpp, daw or egr but could not detect a specific enrichment in cluster C6 similar to our results from PQ-treated S2 cells in vitro (Figure 3B). These findings from the snRNA-seq analysis in vivo and S2 cells in vitro support the hypothesis that cluster C6 plasmatocytes represent a functional state of activated plasmatocytes directly responding to oxidative stress by JNK signaling and Jak/STAT pathway activation. Furthermore, we found evidence that cluster C6 specifically expresses the cytokine upd3, which could reflect an important mediator for oxidative stress response.

Oxidative stress induces distinct cell states in the fat body, including the induction of a Jak/STAT responsive and the loss of an AMP-producing cell cluster

The diversification of plasmatocytes during oxidative stress and the defined cytokine producing immune-activated cell cluster C6 could play a key role in immune response to oxidative stress and also for interorgan communication. To further explore this hypothesis, we aimed to gain further insights into single-cell transcriptomic profiles of the Drosophila fat body during oxidative stress. Using cell type-specific signatures, three of the clusters were annotated as fat body cells (clusters: 1, 3 and 4) (Figure 2A-B). Re-scaling and clustering of all 3150 identified fat body nuclei revealed 8 distinct clusters (Figure 4A). First, we verified fat body-specific signature gene expression across these clusters (Figure S4B) and excluded expression of hemocyte-specific signature genes (Figure S4C). From the eight distinct fat body cell clusters, we identified that cluster C1, C6 and C8 were mainly enriched with cells obtained from control flies, whereas clusters C2-5 were almost exclusively derived from PQ treated flies (Figure 4B, Figure S4D). Only cluster C7 represents cells from both conditions (Figure 4B, Figure S4D). To define the transcriptional changes between these eight fat body cell clusters, we analyzed the TOP20 DE genes across all fat body cell clusters (Figure 4C, Table S2), as well as the unique regulated genes in each specific cluster compared to all others (Figure 4D). Here the cluster C1, mostly derived of fat body cells from control flies, showed an enrichment for gene induction associated with triglyceride storage, such as Fatty acid synthase 1 (FASN1), Glycerol-3-phosphate dehydrogenase 1 (Gpdh1) or Acyl-CoA synthetase long-chain (Acsl), as well as fat body signature genes (Figure 4C-D). Interestingly cluster C8, which mostly consisted of cells from control food, showed a distinct expression of AMP genes including Dro, Drs or Diptericin-B (DptB), however this cluster completely disappeared after PQ feeding (Figure 4C-D). In contrast, PQ treatment induced several fat body specific cell clusters which were not found in control conditions. For example, cluster C2, C4 and C5, which showed a strong gene induction of oxidative stress response genes, but also genes associated with Jak/STAT pathway (e.g. Socs36E, upd3, Fas3, TotA) and JNK activation (e.g. puc, kay) (Figure 4C-D). In all of these clusters, genes associated with triglyceride storage were significantly reduced. This is in line with our findings showing a reduction in triglyceride storage in the fat body upon oxidative stress (Figure 1F, Figure S1E-F). Overall our data revealed that the fat body has a unique heterogeneity upon oxidative stress with a clear separation of clusters with JNK and Jak/STAT activation signature, which could be either activated by the PQ-induced oxidative stress itself or by the cytokines released by immune-activated hemocytes, especially upd3.

Oxidative stress induced different transcriptomic states in fat body cells, including cells with a distinct Jak/STAT activation signature

(A) UMAP visualization of fat body cells from Figure 2A by unsupervised clustering. Colors and dashed lines indicate different clusters. Each dot represents one nucleus. n=3150 nuclei are shown. (B) UMAP of fat body cell clusters split by treatment. Nuclei from control flies are labeled in red and nuclei from PQ treated flies are labeled in blue. Dashed lines indicate different clusters. Each dot represents one nucleus. n=3150 nuclei are shown. (C) Heat map of top 20 DE genes for each fat body cluster. Gene names indicated on the left. Fold change of gene expression is color coded as indicated in legend. (F) Volcano plots comparing pseudo bulk gene expression of individual fat body cluster vs. all fat body cells. The –log10-transformed adjusted P value (P adjusted, y-axis) is plotted against the log2-transformed fold change (FC) in expression between the indicated cluster vs remaining hemocytes (x-axis). Non-regulated genes are shown in grey, not significantly regulated genes are shown in green, significantly regulated genes with a FC<1.5 are shown in blue and significantly regulated genes with a FC>1.5 are shown in red.

Loss of DNA damage signaling in adult hemocytes results in elevated upd3 levels and a decreased survival on oxidative stress

To evaluate the role of immune activation in hemocytes during oxidative stress response, we aimed to specifically explore the implication of DNA damage signaling in hemocytes (Figure 1-3). Previously it has been observed that oxidative stress-induced DNA damage supports immune activation in macrophages via DNA damage signaling27. Based on our findings, we predict that also in adult hemocytes oxidative stress not only drives genome instability and activates repair, but plays a role in the induction and modulation of immune activation via specific gene expression changes in cluster C6 plasmatocytes. To test if DNA damage signaling is involved in the response of adult hemocytes to oxidative stress, we analyzed flies with a hemocyte-specific knock-down for different genes of the DNA damage repair (DDR) machinery including the DNA damage sensing kinases telomere fusion (tefu) and meiotic 41 (mei-41) and the MRN complex protein nbs. Of note, by targeting adult hemocytes, mostly plasmatocytes are targeted which make up 95% of all hemocytes in the adult fly.

We generated Hml>mei-41-IR (w;Hml-Gal4,UAS-2xeGFP/UAS-mei-41-IR), Hml>tefu-IR (w;Hml-Gal4,UAS-2xeGFP/UAS-tefu-IR), Hml>mei41-IR;tefu-IR (w;Hml-Gal4,UAS-2xeGFP/UAS-mei-41-IR;UAS-tefu-IR) and Hml>nbs-IR (w;Hml-Gal4,UAS-2xeGFP/UAS-nbs-IR) flies (Figure S6A). First, we compared their susceptibility to PQ treatment after 18 hrs and found significantly increased susceptibility to the treatment across all DDR-deficient lines compared to the control genotype (Hml/+: 10.7±1.2%, Hml>mei-41-IR: 40.7±4.8%, Hml>tefu-IR: 20.7±4.3%, Hml>mei41-IR;tefu-IR: 37.5±4.8% and Hml>nbs-IR: 48.2±6.7%) (Figure 5A). In these flies, decreased survival after PQ did not correlate to an overall reduction in lifespan on normal food and a decrease in fitness (Figure S6B).

Loss of DNA damage signaling activity in hemocytes leads to an increase in systemic upd3 levels and a higher susceptibility to oxidative stress.

(A) Survival of Hml/+ (n= 28), Hml>mei41-IR (n= 23); Hml>tefu-IR (n= 24); Hml>mei41-IR,tefu-IR (n= 16) and Hml>nbs-IR (n= 22) flies on control food and PQ. Each data point represents a sample of 10 flies. Mean ± SEM is shown. Three independent experiments were performed. One-way ANOVA: ***p<0.001; ****p<0.0001. (B) Hemocyte quantification of Hml/+ control flies and DDR-knockdowns on control food (C) and 15mM Paraquat (P). Each data point represents one fly (n=5 for all groups). Mean ± SEM is shown. One way ANOVA: Hml/+ (B)vs Hml>tefu-IR (C), **p=0.0052; Hml>tefu-IR (C) vs Hml>tefu-IR (P), *p=0.03 (C) Comet assay of isolated hemocytes from Hml/+ (C: n=73; P: n=129), Hml>mei41-IR (C: n=119; P: n=132) and Hml>tefu-IR (C: n=175; P: n=125) - flies on control food and PQ. Olive tail moment of comet assays of sorted hemocyte nuclei is shown. One-way ANOVA: *p<0.05; ***p<0.001; ****p<0.0001. (D) Gene expression analysis for Jak/STAT target genes via RT-qPCR in Hml/+ control flies and DDR-knockdowns on control food (C) and 15mM Paraquat (P). All transcript levels were normalized to the expression of the loading control Rpl1 and are shown in arbitrary units [au]. Each data point (n=5) represents a sample of three individual flies. Mean ± SEM is shown. Two-way ANOVA: *p<0.05; **p<0.01; *** p<0.001; ****p<0.0001. (E) Gene expression analysis for insulin signaling target genes and Ilps via RT-qPCR in Hml/+ control flies and DDR-knockdowns on control food (C) and 15mM Paraquat (P). All transcript levels were normalized to the expression of the loading control Rpl1 and are shown in arbitrary units [au]. Each data point (n=5) represents a sample of three individual flies. Mean ± SEM is shown. Two-way ANOVA: *p<0.05; **p<0.01; *** p<0.001; ****p<0.0001. (F) Triglyceride (TG) levels of Hml/+, Hml>mei41-IR, Hml>tefu-IR, Hml>mei41-IR,tefu-IR and Hml>nbs-IR flies on control food and PQ determined via thin-layer chromatography (TLC). One representative TLC is shown (left panel). Quantification thereof is shown on the right. Each data point (n=4 per group) represents a sample of ten individual flies. Mean ± SEM is shown. n.d. = not detectable. One-way ANOVA was performed for both groups (15mM PQ and control) separately: *p=0.0106.

Next, we inquired if DDR-deficiency in hemocytes might increase susceptibility of the fly to other stresses. For this we challenged all genotypes with starvation. Importantly, we did not observe a decreased survival of DDR-deficient flies on starvation compared to the control genotype, indicating that the observed susceptibility to oxidative stress of flies with a DDR-deficiency in hemocytes is specifically caused by oxidative stress due to PQ treatment (Figure S6C). As defects in the DDR machinery in hemocytes might already affect the development during embryonic and larval stages, we used a hemocyte-specific temperature-inducible knock-down of mei-41, tefu and nbs1 which was induced by a temperature shift to 29°C directly after eclosion from the pupae to induce a knockdown in adult hemocytes only. We analyzed the lifespan of Hml-Gal80ts>mei-41-IR (w;Hml-Gal4,UAS-2xeGFP/UAS-mei-41-IR;tub-Gal80), Hml-Gal80ts>tefu-IR (w;Hml-Gal4,UAS-2xeGFP/UAS-tefu-IR;tub-Gal80), Hml-Gal80ts>mei-41-IR;tefu-IR (w;Hml-Gal4,UAS-2xeGFP/UAS-mei-41-IR;tub-Gal80/UAS-tefu-IR) and Hml-Gal80ts>nbs-IR (w;Hml-Gal4,UAS-2xeGFP/UAS-nbs-IR;tub-Gal80) compared to control flies Hml-Gal80ts/+ (w;Hml-Gal4,UAS-2xeGFP/+;tub-Gal80) and again found no alterations in overall lifespan of all genotypes analyzed on control food except a slight decrease in lifespan of Hml-Gal80ts>mei-41-IR;tefu-IR. (Figure S6D). In contrast, we again found a higher susceptibility to oxidative stress via PQ-feeding in the DDR-deficient lines compared to control (Hml-Gal80ts/+: 30.3±5.2%, Hml-Gal80ts>mei-41-IR: 36.2±4.3%, Hml-Gal80ts>tefu-IR: 52.4±6.9%, Hml-Gal80ts>mei-41-IR;tefu-IR: 72.3±6.3% and Hml Hml-Gal80ts>nbs-IR: 36.2±4.3%) (Figure S6E), which directly confirms our previous results and conclusions (Figure 5A).

As we have seen before that loss of hemocytes is detrimental for the survival on oxidative stress, we wanted to exclude a reduction in hemocytes in the DDR deficient fly lines as a cause of the higher susceptibility to oxidative stress. We quantified GFP+ hemocytes in Hml>mei-41-IR, Hml>tefu-IR, Hml>mei-41-IR;tefu-IR and Hml>nbs-IR flies (Figure 5B, Figure S6A). We did not find alteration in hemocyte number across all fly lines on control food (with the exception of an increase in hemocytes in the Hml>tefu-IR line), nor did we observe any effect of PQ treatment on the number of hemocytes in DDR-deficient flies (Figure 5B, Figure S6A).

Next, we analyzed if the loss of the DDR signaling machinery via knockdown of tefu and mei41 results in increased DNA damage in hemocytes. Comet assays of isolated hemocytes from DDR deficient flies and the control genotype on PQ and control food only indicated a slight increase in DNA damage in these flies, which demonstrates that the higher susceptibility of DDR deficient flies is not due to a loss of hemocyte numbers or a substantial increase in DNA damage, but rather due to changes in the downstream DDR-signaling events modulating immune activation of hemocytes (Figure 5C). Other studies demonstrated that upon tissue damage JNK-mediated secretion of the JAK/STAT cytokines upd1, upd2 and upd3 is a critical step for innate immune signaling28,29. Furthermore, DNA damage signaling is essential to induce such an immune response and to regulate insulin signaling30. To elucidate the implication of DDR signaling on hemocyte activation and oxidative stress resistance, we measured the gene expression levels of upd cytokines as well as JNK and Jak/STAT targets in DDR deficient flies and controls on PQ and control food. Loss of DNA damage signaling in hemocytes induced a significant induction of upd3 expression upon PQ treatment in Hml>tefu-IR and Hml>mei-41-IR;tefu-IR flies (Figure 5D). In contrary, upd1 and upd2 were not significantly changed in DDR-deficient flies (Figure 5D). We measured the Jak/STAT target genes Socs36E and TotA on systemic level. A slight induction of Socs36E expression in all DDR-deficient genotypes on PQ, but only Hml>mei-41-IR;tefu-IR flies showed a significant induction of Socs36E upon PQ treatment compared to control food and Hml/+ flies on PQ (Figure 5D). TotA expression tended to be reduced upon PQ treatment across all genotypes analyzed. However, when we checked the levels of puc as a JNK signaling target gene, we found a significant induction on PQ treatment in Hml>tefu-IR, Hml>mei-41-IR;tefu-IR and Hml>nbs-IR flies, which was not seen in control flies upon PQ treatment (Figure 5D). When we analyzed the gene expression of genes involved in metabolism and energy mobilization, we detected a significant induction and dysregulation of Thor and InR in all DDR-deficient genotypes upon PQ treatment compared to control flies (Figure 5E), as well as a slight induction of Pepck1 (Figure 5E). To exclude that these changes in energy mobilization are due to altered expression of Ilps, we analyzed the levels of Ilp2, Ilp3 and Ilp5 in all genotypes on control and PQ-food, but could not detect remarkable alterations here (Figure 5E). To further exclude that the higher susceptibility of DDR-deficient flies and seen alterations in stress response are due to inefficient mobilization of triglycerides, we measured stored triglycerides across all genotypes and found efficient mobilization of triglyceride storage upon PQ treatment in all genotypes compared to control food (Figure 5F).

From our results we can conclude that DNA damage signaling in adult hemocytes, mostly consisting of plasmatocytes, essentially determines susceptibility to oxidative stress. Loss of DNA damage signaling in hemocytes results in upregulation of JNK signaling and elevated expression of the pro-inflammatory cytokine upd3, as well as transcriptional alterations in energy mobilization and insulin signaling. Based on our previous results on the role of hemocytes during oxidative stress, we can therefore assume that the DNA damage signaling machinery in hemocytes is activated by oxidative stress and plays an essential modulatory and rate-limiting role for subsequent immune activation.

Adult hemocytes control susceptibility to oxidative stress by JNK signaling and upd3 release

To this point, our data suggests that oxidative stress-induced DNA damage signaling in hemocytes limits immune activation and stress response upon oxidative stress. As we found that increased upd3 levels correlate with a higher susceptibility to oxidative stress of flies with DDR-deficient hemocytes, we aimed to test if hemocyte-derived upd3 is a major source to regulate susceptibility to PQ-induced oxidative stress. To do so, we either over-expressed upd3 in hemocytes in Hml>upd3 (w;Hml-Gal4,UAS-2xeGFP/UAS-upd3) flies or performed a hemocyte-specific knockdown of upd3 in Hml>upd3-IR (w;Hml-Gal4,UAS-2xeGFP/UAS-upd3-IR) flies (Figure S6F). Both genotypes showed a normal overall lifespan compared to the Hml/+ control (Figure S6G), but when we compared their susceptibility to PQ-induced oxidative stress, Hml>upd3 flies showed a drastically decreased survival on PQ, whereas Hml>upd3-IR flies did not show a changed survival on PQ treatment after 18 hours (Figure 6A). We tested upd3 expression levels on a whole fly level and found a slight but not significant induction in Hml/+ flies after PQ treatment compared to control food, in line with our previous results (Figure 1B and 6B). As expected, Hml>upd3 flies showed increased levels of upd3 in both treatment groups compared to the other genotypes, but also showed a significant increase of upd3 upon PQ compared to control food (Figure 6B). Surprisingly, Hml>upd3-IR flies still showed a slight induction of upd3 upon PQ treatment, which could potentially be derived from another cellular source after eliminating hemocyte-derived upd3, for example fat body cells, as seen in our snRNA-seq analysis.

Hemocyte-derived upd3 controls susceptibility to oxidative stress in adult Drosophila.

(A) Survival of Hml/+ (n=28), Hml>upd3-IR (n=27) and Hml>upd3 (n=26) flies on 15mM PQ food. Each data point represents a sample of 10-15 flies. Mean ± SEM is shown. Three independent experiments were performed. One-way ANOVA: ****p<0.0001. (B) Gene expression analysis for upd3 via RT-qPCR in Hml/+, Hml>upd3-IR and Hml>upd3 flies on control food (C) and 15mM Paraquat (P). Each data point (n=5-6) represents a sample of three individual flies. Mean ± SEM is shown. Two-way ANOVA: ****p<0.0001. (C) Hemocyte quantifications in Hml/+, Hml>upd3-IR and Hml>upd3 flies on control food and 15mM PQ food. Each data point represents one fly (n= 5-6). Mean ± SEM is shown. One-way ANOVA: **p<0.01. (D) Survival of Hml/+ (n=241), upd3null;Hml/+ (n=122), Hml>upd3 (n=104), upd3null;Hml>upd3 (n=57), Hml>mei-41-IR;tefu-IR (n=103) and upd3null;Hml>mei-41-IR;tefu-IR (n=50) flies on 15mM PQ food at 29°C. Each data point represents a sample with 10-15 flies. Mean ± SEM is shown. One-way ANOVA: ****p<0.0001. (E) Survival of Hml/+ (n=78), Hml>hep[act] (n=38) and Hml>bsk-DN (n=80) flies on 15mM PQ food. Each data point represents a sample with 10 flies. Mean ± SEM is shown. One-way ANOVA: **p<0.01; ****p<0.0001. (F) Gene expression analysis for upd3 via RT-qPCR of Hml/+, Hml>hep[act] and Hml>bsk-DN flies on control food (-) and 15mM Paraquat (+). Each data point (n=5-6) represents a sample of three individual flies. Mean ± SEM is shown. Two-way ANOVA: *p<0.05.

To identify if the higher susceptibility of Hml>upd3 flies to oxidative stress was due to reduced hemocyte numbers, we quantified hemocytes across all three genotypes, but did not see any significant changes of hemocyte number in both genotypes compared to Hml/+ control flies on control food and PQ (Figure 6C and Figure S6F). Only in the Hml>upd3-IR flies we found a reduction of hemocytes between flies on control food and PQ, but without consequences for their susceptibility to oxidative stress (Figure 6C and Figure S6G). Our data indicate that elevation of hemocyte-derived upd3 is sufficient to increase the mortality of adult flies on PQ-mediated oxidative stress. Yet we could not exclude other cellular sources than hemocytes for upd3 upon oxidative stress. Therefore, we wanted to probe if hemocyte-derived upd3 is sufficient to increase the susceptibility to PQ treatment in the absence of other cellular sources. We backcrossed Hml>upd3 flies to upd3-null mutants (upd3null) to generate upd3null;Hml>upd3 flies (upd3null;Hml-Gal4,UAS-2xeGFP/UAS-upd3) and compared their survival on PQ to Hml/+ and upd3null flies (Figure 6D). Of note, we found that upd3nullmutants had a significant reduced susceptibility to oxidative stress compared to Hml/+ flies, however we found a decreased survival in upd3null;Hml>upd3 flies when hemocytes alone overexpressing upd3 similar to the higher susceptibility of Hml>upd3 flies (Figure 6D). We then wanted to test if loss of upd3 would also rescue the higher susceptibility of flies with DDR-deficient hemocytes and backcrossed Hml>mei41-IR;tefu-IR to upd3nullflies to generate upd3null;Hml>mei41-IR;tefu-IR flies (upd3null;Hml-Gal4,UAS-2xeGFP/UAS-mei41-IR;UAS-tefu-IR) (Figure 6D). Indeed we found a substantially decreased susceptibility of these flies to oxidative stress compared to Hml>mei41-IR;tefu-IR flies, supporting our hypothesis that loss of DNA damage signaling in hemocytes and subsequently increased levels of upd3 render adult flies more susceptible to oxidative stress.

Previous studies highlighted that JNK signaling can induce upd3 expression in hemocytes12. In line with these reports, we found enrichment of genes with kayak/AP-1 binding motifs in the plasmatocyte subset C6 which showed further upregulation of Jak/STAT target genes and also upd3 expression (Figure 3). Therefore, we wanted to verify if JNK activation in hemocytes acts as a direct downstream signal activated by oxidative stress and subsequently induces the release of the pro-inflammatory cytokine upd3. We either overexpressed a constitutive active form of JNKK hemipterous (hep) in hemocytes using Hml>hep[act] (w;Hml-Gal4,UAS-2xeGFP/UAS-hep[act]) flies or overexpressed a dominant-negative version of the Jun kinase basket (bsk) in hemocytes using Hml>bsk-DN (w;Hml-Gal4,UAS-2xeGFP/UAS-bsk-DN). We compared their survival on PQ after 18 hrs and in line with our previous results, we found an increased susceptibility of Hml>hep[act] flies to PQ treatment, but not of Hml>bsk-DN flies compared to the control genotype (Figure 6E). We again analyzed the overall lifespan of these flies on normal food and could not find differences between the genotypes (Figure S6H). Furthermore, when we analyzed upd3 expression in these flies upon PQ treatment, we found a significant induction of upd3 expression after PQ treatment in Hml>hep[act] compared to control food, but only a slight increase in Hml/+ and Hml>bsk-DN flies upon PQ feeding (Figure 6F). These findings support a substantial role of JNK-mediated induction of upd3 in hemocytes upon oxidative stress.

In summary, our study revealed a defined immune-responsive subset of plasmatocytes during oxidative stress. We show that this oxidative stress-mediated immune activation seems to be regulated by DNA damage signaling in hemocytes, which controls JNK activation and subsequent upd3 expression. Loss of DNA damage signaling and over-induction of JNK signaling or upd3 in hemocytes release renders the adult fly more susceptible to oxidative stress.

Discussion

Similar to vertebrate macrophages, oxidative stress triggers immune activation in Drosophila hemocytes18. However, this immune activation was not obvious on a systemic level upon PQ-mediated oxidative stress. In contrast, localized oxidative stress-induced immune activation in damaged tissues is an essential step of the Drosophila immune response to infection or injury19,31. However, the precise contribution of Drosophila macrophages in this immune activation, in the organ-to-organ communication during stress response and the implication in the organism’s survival upon increased oxidative stress remained so far ill-defined. Overall, we describe the induction of diverse cellular states of adult hemocytes, namely plasmatocytes, during oxidative stress with a particular subset of plasmatocytes (cluster C6) which directly responds to oxidative stress by immune activation. We demonstrated that this immune activation seems to be at least partially controlled by the DNA damage signaling machinery in hemocytes. Loss of nbs, tefu or mei-41 in hemocytes renders the organism more susceptible to oxidative stress without significantly affecting oxidative stress mediated DNA damage, but with an enhanced induction of JAK/STAT and JNK signaling, accompanied by changes in genes associated with energy mobilization and storage. In line with that, we describe substantial changes in the fat body on the transcriptomic level with diverse cell states, indicating decrease of triglyceride storage and induced response to oxidative stress and immune activation, but loss of AMP-producing cells. Finally, we showed that the control of hemocyte-derived upd3 is key for the survival of the organism upon oxidative stress and that induction of the JNK/upd3 signaling axis in hemocytes is sufficient to increase the susceptibility to oxidative stress. Hence, we postulate that hemocyte-derived upd3 is most likely released by the activated plasmatocyte cluster C6 during oxidative stress in vivo and is subsequently controlling energy mobilization and subsequent tissue wasting upon oxidative stress.

Drosophila hemocytes have been described before to react to PQ-mediated or infection-mediated intestinal damage via release of cytokines to promote ISC proliferation in the gut14,17. PQ is a widely used model in Drosophila to induce oxidative stress via oral uptake32. In many reports the gut has been described to be the most affected organ and ISC proliferation is an essential step to regenerate the injured intestinal epithelium33,34. PQ treatment has been described to cause gut leakage and therefore this could be a potential cause of death for the fly14, though we did not find signs of gut leakage within the 18 hours PQ treatment with a concentration of 15mM. Aside from the gut, oral PQ treatment was shown to induce oxidative stress in many organs including hemocytes35, oenocytes36 and the brain37, indicating that the induced oxidative stress is not only localized to the gut tissue. PQ-induced oxidative stress was described before to cause neurodegeneration with parkinsonism-like symptoms, accompanied by strong immune activation and upregulation of JNK signaling in the CNS37, an activation profile similar to the identified activated plasmatocyte cluster C6 and the activated fat body clusters C2 and C4 in our snRNA-seq analysis. The snRNA-seq analysis further revealed distinct hemocyte clusters with seven plasmatocyte clusters and one crystal cell cluster. Whereas the plasmatocyte clusters C1 and C7 seem to represent cellular states found in steady state conditions only, we identified several other clusters which seemed to be exclusive or at least enriched under PQ treatment. Interestingly, we only identified one cluster, C6, which specifically responded to oxidative stress via immune activation and upregulation of oxidative stress response genes but also Jak/STAT and JNK target genes. We postulate that these cells might be directly exposed to the oxidative stress because their signature resembles the one of S2 cells directly treated with PQ in vitro. Even though oxidative DNA damage in hemocytes was followed by immune activation in vitro and in vivo, it remains to be determined if this activation is induced by cytokines derived from damaged organs during oxidative stress or by the oxidative stress directly. The hypothesis that this cluster is directly exposed to oxidative stress is further supported by an upregulation of genes such as Hsps in this cluster which are known to be directly induced by oxidative stress38. Furthermore, we found evidence that oxidative DNA damage and induced DNA damage signaling dynamically modulate immune activation cascades in cluster C6 plasmatocytes. This was supported by the increased DNA damage in S2 cells by PQ treatment but also the enrichment of DDR-associated genes such as Gadd45 in cluster C6, which is associated with JNK activation and is further implicated in oxidative stress induced DNA damage37. Constant DNA damage signaling and repair is an essential step in all tissues to maintain tissue homeostasis39.

In our study, we point to a potential role of DNA damage signaling as a rate-limiting step for immune activation and response to oxidative stress in hemocytes. We did not find indications of canonical DNA damage response, such as induced apoptosis in hemocytes due to the oxidative DNA damage. Non-canonical DNA damage sensing has been implicated in immune responses in vertebrates and invertebrates30,37,40,41. It was demonstrated that vertebrate macrophages in vitro respond similarly to immune stimuli such as infection by synergistic activation of DDR and type I interferon signaling to induce downstream immune signaling cascades27. Here, we provide new evidence in Drosophila that potentially non-canonical DNA damage is used by immune cells in vivo to modulate and especially control their immune response including pro-inflammatory cytokine release upon oxidative stress. Localized DNA damage in larval epithelial cells has been shown to induce non-canonical DNA damage activity leading to the secretion of upds and the regulation of hemocyte expansion and activation, as well as subsequent hemocyte-mediated regulation of energy storage in the fat body30. In further support of our results, one report previously demonstrated that loss of DNA damage signaling machinery induced the release of upd3 in hemocytes14. Most likely, the identified plasmatocyte cluster C6 is a central cellular source of this signal. However, we also detected upd3 induction in the PQ-induced fat body cell cluster C4 as potential secondary source of upd3, which needs to be further determined in future studies. Furthermore, anatomical localization of this upd3-producing plasmatocyte cluster, as well as spatial distribution and cross-talk of the different plasmatocyte and fat body cell clusters would be of interest. We describe a strong connection between the control of upd3 expression by hemocytes and the susceptibility of the adult fly to oxidative stress. Upon oxidative stress, we observe decreased triglycerides and glycogen stores and altered gene expression such as induced expression of lipolytic enzymes and other foxo target genes indicating energy mobilization from the adult fat body, which was further supported by the transcriptomic profiling of fat body cells. Accumulating evidence throughout our study points to an implication of hemocyte-derived upd3 in the control of energy mobilization and tissue wasting during oxidative stress. Our data from hemocyte-deficient flies indicating defects in energy mobilization and maintenance of free glucose levels further support this assumption. Our presented findings are in line with our previous studies where we showed that upon high fat diet hemocytes produce high levels of upd3 which induces insulin insensitivity in many tissues by activating Jak/STAT signaling in these organs and inhibiting the insulin signaling pathway12. Furthermore, we and others described a central role for hemocyte derived upds in the control of glucose metabolism in muscles during steady state and upon infection13,42. As shown before, hemocytes produce high levels of upd3 upon infection with Mycobacterium marinum which also caused tissue wasting and extensive energy mobilization in a foxo-dependent manner11,43.

Our study provides further evidence that hemocytes serve as essential signaling hubs in “organ to organ communication” during oxidative stress, which is in line with previous studies in immune challenge as well as steady state19,42,44. Our findings in Drosophila highlight the role of macrophages for the survival on oxidative stress with an urgent need for a balanced immune response to overcome the tissue damage. We further unravel the diverse responses of hemocytes and fat body cells towards oxidative stress which can now serve as a baseline to further explore the spatiotemporal macrophage/hemocyte responses to oxidative stress and their communication with other host tissue cells such as the fat body. Future in depth analysis of the regulation of these cellular states and their coupling to tissue repair will allow to gain new insights for the treatment of diseases where tissue repair and tissue wasting needs to be balanced, including cancer or neurodegenerative diseases.

Supporting information